---
title: 'Analyzing and Predicting Survival on Titanic'
author: 'Prasad Sawant'
date: '26 August 2016'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---
# Introduction
This is my first attempt using R for Data Analysis and Prediction

## Load packages

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('dplyr') # data manipulation
library('rpart') # prediction model


```
## Load data

```{r, message=FALSE, warning=FALSE}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)
full  <- bind_rows(train, test) # bind training & test data

```

# Feature Engineering

Let's explore a few variables to see their correlation with Survival

## Age

```{r, message=FALSE, warning=FALSE}
ggplot(full[1:891,], aes(x=Age, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge')

```

We can see a pattern that for some age groups survival rate is higher. But not conclusive as too many age values are present. 
So lets create age groups where there are only few groups to explore.

```{r, message=FALSE, warning=FALSE}

full$AgeGroup[full$Age > 0 & full$Age <=10] <- '0-10'
full$AgeGroup[full$Age > 10 & full$Age <=20] <- '10-20'
full$AgeGroup[full$Age > 20 & full$Age <=30] <- '20-30'
full$AgeGroup[full$Age > 30 & full$Age <=40] <- '30-40'
full$AgeGroup[full$Age > 40 & full$Age <=50] <- '40-50'
full$AgeGroup[full$Age > 50 & full$Age <=60] <- '50-60'
full$AgeGroup[full$Age > 60 & full$Age <=70] <- '60-70'
full$AgeGroup[full$Age > 70 & full$Age <=80] <- '70-80'
full$AgeGroup[full$Age > 80 & full$Age <=90] <- '80-90'
full$AgeGroup[full$Age > 90 & full$Age <=100] <- '90-100'
full$AgeGroup[full$Age == ''] <- 'Unknown'

ggplot(full[1:891,], aes(x=AgeGroup, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge')

```

Couple of things to note,

1) The chances of survival are more if you are a child (0-10 years)
2) There are quite a lot missing values of Age

Let us broadly classify the Age in 2 categories as Children and Adults. Also lets derive the AgeType from the Title present in the Name.

```{r, message=FALSE, warning=FALSE}
full$AgeType[full$Age > 0 & full$Age <=10] <- 'Child'
full$AgeType[full$Age > 10 & full$Age <=120] <- 'Adult'

full$AgeType[(is.na(full$Age)) & grepl('Master.', full$Name)] <- 'Child'
full$AgeType[(is.na(full$Age)) & grepl('Miss.', full$Name)&(full$Parch>0)] <- 'Child'
full$AgeType[(is.na(full$Age)) & grepl('Miss.', full$Name)&(full$Parch==0)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Ms.', full$Name)&(full$Parch>0)] <- 'Child'
full$AgeType[(is.na(full$Age)) & grepl('Ms.', full$Name)&(full$Parch==0)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Mlle.', full$Name)&(full$Parch>0)] <- 'Child'
full$AgeType[(is.na(full$Age)) & grepl('Mlle.', full$Name)&(full$Parch==0)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Mrs.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Mr.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Dr.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Rev.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Mme.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Major.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Col.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Capt.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Dona.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Jonkheer.', full$Name)] <- 'Adult'
full$AgeType[(is.na(full$Age)) & grepl('Countess.', full$Name)] <- 'Adult'


ggplot(full[1:891,], aes(x=AgeType, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge')
```

Now lets see how does being aboard with Family affect Survival

## Family member

Although this attribute does not exist but it can be derived from 'SibSp' and 'Parch'

```{r, message=FALSE, warning=FALSE}
full$familySize <- full$SibSp + full$Parch

```
Now lets see the correlation of FamilySize to Survival

```{r, message=FALSE, warning=FALSE}
ggplot(full[1:891,], aes(x=familySize, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge') +
scale_x_continuous(breaks=c(1:11)) 

missFamilySize <- sum(is.na(full$familySize))
missFamilySize

```

Looks like people who are travelling alone have less chances of survival
However people travelling with family size between 1-4 have good chances of survival
Survival rate again goes down once family size exceeds 4


## Pclass
```{r, message=FALSE, warning=FALSE}

ggplot(full[1:891,], aes(x=Pclass, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge') +
scale_x_continuous(breaks=c(1:11)) 

```

Seems like people travelling on Passenger class 1 have more chances of survival. While people travelling on Passenger class 3 have lower chances of Survival

## Embarkment
```{r, message=FALSE, warning=FALSE}

missEmb <- sum(is.na(full$Embarked))
missEmb

full$Name[is.na(full$Embarked)]

ggplot(full[1:891,], aes(x=Embarked, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge')
```

Passengers with Embarkment of type 'S' show lower chances of survival

## Gender

```{r, message=FALSE, warning=FALSE}


missGen <- sum(is.na(full$Sex))
missGen

ggplot(full[1:891,], aes(x=Sex, fill=factor(Survived))) +
geom_bar(stat='count', position='dodge')

```

Also female passengers have higher chances of survival corresponding to male passengers


# Create Prediction Model
Let us use the decision tree model for predicting the surival


```{r, message=FALSE, warning=FALSE}
train <- full[1:891,]
test <- full[892:1309,]

survival_model <- rpart(factor(Survived) ~ AgeType + familySize + Pclass + Embarked + Sex, data = train)
plot(survival_model, margin=0.1)
text(survival_model, use.n=TRUE, pretty=TRUE, cex=0.8)

predict_survival <- predict(survival_model, test, type = "class")

solution <- data.frame(PassengerID = test$PassengerId, Survived = predict_survival)
write.csv(solution, file = 'survival_model_solution.csv', row.names = F)


```


