---
output:
  html_document: default
  pdf_document: default
---

# EDA on the Titanic Dataset
  
### In this Kernel, I have prepared a simple Exploratory Analysis to understand details of the Titanic dataset and visualize the characteristics of its variables, hoping this will help later with the creation and evaluation of predictive models.

### I have divided this kernel into three parts:
  
* ### Feature Engineering
* ### Studying and Treating Missing Values
* ### Exploratory Analysis
  
### Let's load the packages that we need for this analysis and load our data
```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(flexdashboard)
library(ggplot2)
library(readr)
library(highcharter)
library(dplyr)
library(bindrcpp)
library(gridExtra)
library(rpart)
library(corrplot)
library(randomForest)

train <- read.csv('../input/train.csv')
test <- read.csv('../input/test.csv')
total <- rbind(train[,-2],test)

# Keeping a version of the original dataset, will be used later
total2 <- total
```

## Feature Engineering
#### We can break the variable "Name" into the name itself and title. This title can add meaningful information into our analysis since it depicts certain categories of passengers.

### Grab title from passenger names
```{r}
total$Title <- gsub('(.*, )|(\\..*)', '', total$Name)
```
  
#### Titles with very low cell counts to be combined to "rare" level
```{r}
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')
```
```{r}
# Also reassign mlle, ms, and mme accordingly
total$Title[total$Title == 'Mlle']        <- 'Miss' 
total$Title[total$Title == 'Ms']          <- 'Miss'
total$Title[total$Title == 'Mme']         <- 'Mrs' 
total$Title[total$Title %in% rare_title]  <- 'Rare Title'
total$Title <- factor(total$Title)
# Show title counts by sex again
# table(train$Sex, train$Title)
```
```{r}
# Finally, grab surname from passenger name
total$Surname <- sapply(as.character(total$Name),  
                        function(x) strsplit(x, split = '[,.]')[[1]][1])

total$Surname <- factor(total$Surname)
```

### Adding Mother variable
```{r}
total$Mother <- 'Not Mother'
total$Mother[total$Sex == 'female' & total$Parch > 0 & total$Age > 18 & total$Title != 'Miss'] <- 'Mother'
total$Mother <- factor(total$Mother)
```

```{r}
train <- train[train$Embarked!='',]
train$Embarked <- factor(train$Embarked, labels = c("Cherbourg", "Queenstown", "Southhampton"))
```
## ____________________________________________
## Studying the NAs

### Now we analyze the number of missing values for the remaining variables
```{r}
Num_NAtrain<-sapply(train[,-2],function(y)length(which(is.na(y)==T)))
Train_Empty<-sapply(train[,-2],function(y)length(which(y=="")))
Num_NAtest<-sapply(test,function(y)length(which(is.na(y)==T)))
Test_Empty<-sapply(test,function(y)length(which(y=="")))
NA_Count<- data.frame(NATrain=Num_NAtrain,NATest=Num_NAtest,Empty_Train = Train_Empty, Test_Empty = Test_Empty,
                      Sum_NA=Num_NAtrain+Num_NAtest,Sum_Empty=Train_Empty+Test_Empty)
NA_Count
```
  
    
#### The variable "Age" has 20% of missing values in the train dataset and 21% of missing values in the test data set. This is an important predictor that carries relevant information for any future models we might create. We cannot ignore it simply because it has many missings. We need to find a way to address this problem later. The variable "Fare", however, has only one NA so we can remove that row.
  
  
#### The variable "Cabin" is composed mostly of empty cells, so we will drop that variable. 
  
### I will create another variable based on the predictor "Ticket". This new variable will take only the first four characters of the variable "Ticket"
### After this, I will transform this new variable into numerical type. By doing this, I am trying to see if this variable can be used in any way to help us later with prediction.
```{r}
# Is there any information we can obtain from the variable "Ticket"?
total$ticketclass <- substr(total$Ticket, 1, 4)
total$ticketclass <- as.numeric(as.integer(factor(total$ticketclass)))
```
  
### Creating rpart and random forest models to fill NAs
```{r}
data1 <- total[!is.na(total$Age),-c(1,3,8,10,13)]
data2 <- total[is.na(total$Age),]
data1train <- data1[1:800,]
data1test <- data1[-c(1:800),]
model <- rpart(Age~., data1train,control = rpart.control(cp = 0.01, minsplit = 30, xval = 700,maxdepth = 20, minbucket = 1))
modelrf <- randomForest(Age~., data1train, ntree = 300, nodesize=1)
data2$Age <- NULL
data3 <- data2
data2$Age <- predict(model,data2[,-c(1,3,7,9)])
data3$Age <- predict(modelrf,data2[,-c(1,3,7,9)])
data1 <- total[!is.na(total$Age),]
total_merged <- rbind(data1,data2)
total_merged2 <- rbind(data1,data3)
```
  
#### Let's see how the models have changed the distribution of the variable "Age"
  
```{r, echo=F}
par(mfrow=c(1,3))
hist(total$Age, freq=F, main='Age: Original Data', 
     col='yellow', ylim=c(0,0.05))
hist(total_merged$Age, freq=F, main='Age: Rpart output', 
     col='darkgreen', ylim=c(0,0.05))
hist(total_merged2$Age, freq=F, main='Age: Random Forest output', 
     col='lightgreen', ylim=c(0,0.05))
```
  
### So the dataset "total_merged2" will be now our "total" dataset
  
```{r, include=F}
total <- total_merged2

total <- total[order(total$PassengerId),]
```

### Let's reassemble our train dataset
```{r}
train_new <- total[1:889,]
train <- cbind(train_new,Survived = train$Survived)
```
## ____________________________________________
## Now we can start our Exploratory Analysis
  
### First, let's see what filling the NAs has done to our "Age" variable, by comparing the previous distribution with the NAs against the distribution that had the NAs filled:
  
### Age Original
```{r, echo = F}

tmp_male <- total2 %>%  filter(Sex=="male", !is.na(Age)) %>% select(Age) %>% .[[1]]
b <- hist(tmp_male, 20, plot=FALSE)
tmp_female <- total2 %>%  filter(Sex=="female", !is.na(Age)) %>% select(Age) %>% .[[1]]
a <- hist(tmp_female, breaks = b$breaks, plot=FALSE)

df <- data.frame(Age=c(a$mids,b$mids),Density=c(a$density,b$density),Sex=c(rep("female",length(a$mids)),rep("male",length(b$mids))))

highchart() %>% 
  hc_add_series(name="female", select(filter(df,Sex=="female"),Density)[[1]], type="column", color='rgba(255, 192, 203, 0.30)', showInLegend=FALSE) %>% 
  hc_add_series(name="male", select(filter(df,Sex=="male"),Density)[[1]], type="column", color='rgba(68, 170, 255, 0.30)', showInLegend=FALSE) %>% 
  hc_add_series(name="male", select(filter(df,Sex=="male"),Density)[[1]], type="spline", color="#44AAFF") %>% 
  hc_add_series(name="female", select(filter(df,Sex=="female"),Density)[[1]], type="spline", color="#FFC0Cb") %>% 
  hc_tooltip(pointFormat = "<span style=\"color:{series.color}\">{series.name}</span>:
             {point.y:.3f}<br/>",
             shared = FALSE) %>% 
  hc_yAxis(title=list(text='Density')) %>% 
  hc_xAxis(title=list(text='Age'))   
```

### Age Modified
```{r, echo = F}

tmp_male <- total %>%  filter(Sex=="male", !is.na(Age)) %>% select(Age) %>% .[[1]]
b <- hist(tmp_male, 20, plot=FALSE)
tmp_female <- total %>%  filter(Sex=="female", !is.na(Age)) %>% select(Age) %>% .[[1]]
a <- hist(tmp_female, breaks = b$breaks, plot=FALSE)

df <- data.frame(Age=c(a$mids,b$mids),Density=c(a$density,b$density),Sex=c(rep("female",length(a$mids)),rep("male",length(b$mids))))

highchart() %>% 
  hc_add_series(name="female", select(filter(df,Sex=="female"),Density)[[1]], type="column", color='rgba(255, 192, 203, 0.30)', showInLegend=FALSE) %>% 
  hc_add_series(name="male", select(filter(df,Sex=="male"),Density)[[1]], type="column", color='rgba(68, 170, 255, 0.30)', showInLegend=FALSE) %>% 
  hc_add_series(name="male", select(filter(df,Sex=="male"),Density)[[1]], type="spline", color="#44AAFF") %>% 
  hc_add_series(name="female", select(filter(df,Sex=="female"),Density)[[1]], type="spline", color="#FFC0Cb") %>% 
  hc_tooltip(pointFormat = "<span style=\"color:{series.color}\">{series.name}</span>:
             {point.y:.3f}<br/>",
             shared = FALSE) %>% 
  hc_yAxis(title=list(text='Density')) %>% 
  hc_xAxis(title=list(text='Age'))   
```
  
#### We can see that filling the NAs has greatly affected the distribution of the variable, even though that change did not seem so significant before. Since we are just doing Exploratory Analysis, we will stop here, but this issue should be considered if we are proceeding with creating predictive models.

### Let's proceed with our study of the variables
  
### Survival rates of titles
```{r, tidy=T, echo=F}
tmp <- train %>%  group_by(Title) %>% summarize(Survived = mean(Survived))
tmp$colors <- c("#d35400", "#2980b9", "#2ecc71", '#e4d354', '#2b908f')
hchart(tmp, "column", hcaes(x = Title, y = Survived, color=colors)) %>% 
  hc_tooltip(pointFormat = "{point.y:.2f}</br>",shared = FALSE)

```
  
### Port of Embarkation
```{r, echo=F}

tmp <- train %>% filter(!(Embarked=="")) %>% group_by(Embarked) %>% tally() %>% mutate(Percent = n/sum(n))
tmp$colors <- c("#d35400", "#2980b9", "#2ecc71")
tmp <- arrange(tmp,desc(Percent))
highchart() %>% 
  hc_xAxis(categories = c("Southhampton", "Cherbourg", "Queenstown")) %>%
  hc_yAxis(title=list(text='Percentage')) %>%
  hc_add_series(tmp, "bar", hcaes(x = Embarked, y = Percent, color=colors)) %>% 
  hc_tooltip(pointFormat = "{point.y:.2f}</br>",shared = FALSE) %>% 
  hc_legend(enabled=FALSE)

```
  
### Passenger Class
```{r, echo=F}

tmp <- train %>%  group_by(Pclass) %>% summarize(Survived = mean(Survived))
tmp$colors <- c("#d35400", "#2980b9", "#2ecc71")
hchart(tmp, "column", hcaes(x = Pclass, y = Survived, color=colors)) %>%  
  hc_tooltip(pointFormat = "{point.y:.2f}</br>",shared = FALSE)

```
  
### Embark
```{r, echo=F}

tmp <- train %>% filter(!(Embarked=="")) %>% group_by(Embarked) %>% summarize(Survived = mean(Survived))
tmp$colors <- c("#d35400", "#2980b9", "#2ecc71")
hchart(tmp, "column", hcaes(x = Embarked, y = Survived, color=colors))
```
  
### Sex
```{r, echo=F}
tmp <- train %>% group_by(Sex) %>% tally() %>% mutate(pct = n/sum(n))
tmp$colors <- c('#7cb5ec', '#434348')

hchart(tmp, "pie", hcaes(x = Sex, y = pct, color=colors))
```
  
### Parch
```{r, echo=F}

train$Parch = factor(train$Parch)

tmp <- train %>% group_by(Parch) %>% tally() %>% mutate(pct = n/sum(n))
tmp$colors <- c('#7cb5ec', '#434348', '#90ed7d', '#f7a35c', '#8085e9', 
   '#f15c80', '#e4d354')

hchart(tmp, "pie", hcaes(x = Parch, y = pct, color=colors))
```
  
### Plot age distributions
```{r, echo=F}


par(mfrow=c(1,2))
hist(train$Age, freq=T, main='Income: Train Data', 
     col='darkgreen', ylim=c(0,55),breaks = 50)
hist(test$Age, freq=T, main='Income: Test Data', 
     col='lightgreen', ylim=c(0,55),breaks = 50)
```

### Out of curiosity, let's analyze the variable "Fare" vs. "Sex" with "Pclass" as filter
```{r, echo=F, warning=F}
ggplot(total, aes(x=Pclass,y=Fare,fill=Fare,colour=Fare)) + 
  geom_bar(stat='identity') + facet_grid(Sex ~ .)
```
  
### Let's see what was the survival rate for the classes of the variable "Mother"
```{r, echo=F}
tmp <- train %>%  group_by(Mother) %>% summarize(Survived = mean(Survived))
tmp$colors <- c("#d35400", "#2980b9")
hchart(tmp, "column", hcaes(x = Mother, y = Survived, color=colors)) %>% 
  hc_tooltip(pointFormat = "{point.y:.2f}</br>",shared = FALSE)
```
  
### Finally, let's study the correlations among the variables

```{r, echo = F}
#total1 <- total
par(mfrow=c(1,1))
for(i in 1:14){
  if(is.factor(total[,i])){
    total[,i]<-(as.integer(total[,i]))
  }
}

total[,-1] %>%
  cor(use="complete.obs", method="spearman") %>%
  corrplot(type="lower", method="number", diag=FALSE)
```

### ...
### To be continued

