# Introduction

This script analyzes the expected score if only 
genders are considered and no other external or cross information.
are taken into account.

The next plots show there were twice more males aboard than females.
Also, female survival rate was more than twice higher than for males.
 
Upon these observations, the optimal prediction is 
all females survive and all males perish.

The expected score by processing the  training set is 78.60%.
The actual obtained score  on the public set is 76.56% (You'll get 0.7655 points)

Since this solution is optimal within the constraint, it implies that cross validation 
during other optimization methods is potentially close to 2% higher than the leader board 
score when Gender becomes a major considered feature.

```{r, message=TRUE, warning=FALSE, include=FALSE}
rm(list = ls());
library(ggplot2);
library(gridExtra);
library(dplyr);
path <- "../data/";
if(!dir.exists(path)) {
  path <- "../input/"; # changing path to Kaggle's environment
}
```

```{r, echo=FALSE, message=TRUE, warning=FALSE}
# read raw data
load_data <- function(file) {
	return (read.csv(paste0(path, file)));
};

train <- load_data("train.csv");
test  <- load_data("test.csv");
```

```{r, message=TRUE, warning=FALSE, include=FALSE}
# combine test and train for pre-processing
test$Survived <- rep(NA, nrow(test));
comb <- rbind(train, test);
```

```{r, echo=TRUE, message=TRUE, warning=FALSE}
# get index list for future re-split
train_index <- comb$PassengerId[!is.na(comb$Survived)] 
test_index  <- comb$PassengerId[ is.na(comb$Survived)] 
comb$Set <- ifelse(comb$PassengerId %in% train_index, "Train", "Test");
```

This graph essentially says that if one considers only gender as a feature, the optimum strategy for prediction is to predict all females survive and all male perish.

```{r, echo=TRUE, message=TRUE, warning=FALSE}
# plot survival per gender in train set

dftotal    <- comb %>% filter(Survived==0| Survived==1) %>% select(Survived, Sex, Set) 
dfsurvived <- comb %>% filter(Survived==1) %>% select(Survived, Sex, Set) 
dftotal$count    <- "Total"
dfsurvived$count <- "Survived"
df <- rbind(dftotal, dfsurvived);

ggplot(df, aes(Sex, fill = count)) + geom_bar(position="dodge") + labs(title="Survival per Gender in Train Set") + xlab("Gender")
```


The section below investigates a discrepancy between the opening problem statement 
and the train set.
So far other test seems to indicate this discrepancy can be ignored.
In other words, the survival rate in train/test set is higher than the
the ratio in the opening statement.

```{r, message=TRUE, warning=FALSE, include=FALSE}
total  <- 2224;
killed <- 1502;
```

According to the opening problem statement `r killed` perished out of `r total` passangers.

```{r, message=TRUE, warning=FALSE, include=FALSE}
true_survival_rate <- 100*(total-killed)/total;
```

The true (opening statement) survival rate `r  true_survival_rate` percent.

```{r, message=TRUE, warning=FALSE, include=FALSE}
pop_count       <- comb %>% filter(Set=="Train") %>% count(Sex);
pop_surv_count  <- comb %>% filter(Set=="Train" & Survived==1) %>% count(Sex); 
surv_likelihood <- pop_surv_count$n/pop_count$n;
comb$Prediction <- ifelse(surv_likelihood[comb$Sex] > 0.5, 1, 0);
```

Estimated score from train set `r sum(comb$Survived==comb$Prediction, na.rm=T)/nrow(train)`

```{r, message=TRUE, warning=FALSE, include=FALSE}
# Train set survival rate
train_survival_rate <- 100*sum(pop_surv_count$n)/sum(pop_count$n);
print(sprintf("Train set survival rate %2.1f percent", train_survival_rate));
print(sprintf("Survival rate overestimation in train set = %2.1f percent", train_survival_rate - true_survival_rate));
```

#                               Train set statistics

```{r, echo=FALSE, message=TRUE, warning=FALSE}
map.gen <- c("female"=1, "male"=2);
weight <- ifelse( comb$Set=="Train", 100/length(train_index), 100/length(test_index));
ggplot(comb, aes( x = Sex, fill=Set)) + geom_bar(position="dodge", aes(weight=weight)) +
  labs(title="Gender Distribution in Train and Test Sets", x="Gender", y="Percent");

print(sprintf("Female to population ratio in train set = %2.1f percent", 100*pop_count$n[map.gen["female"]]/sum(pop_count$n)));
print(sprintf("Male   to population ratio in train set = %2.1f percent", 100*pop_count$n[map.gen["male"]]/sum(pop_count$n)));

train_gender_survived <- table(train$Sex[train$Survived==1]);

print(sprintf("Female survival rate in train set = %2.1f percent", 100*pop_surv_count$n[map.gen["female"]]/pop_count$n[map.gen["female"]]));
print(sprintf("Male   survival rate in train set = %2.1f percent", 100*pop_surv_count$n[map.gen["male"]]/pop_count$n[map.gen["female"]]));
```

#                              Test set statistics

```{r, echo=FALSE, message=TRUE, warning=FALSE}
test_pop_count   <- comb[test_index,]  %>% count(Sex) 
# calculate likelihood of survival per sex

print(sprintf("Female to population ratio in test set = %2.1f percent", 100*test_pop_count$n[map.gen["female"]]/sum(test_pop_count$n)));
print(sprintf("Male   to population ratio in test set = %2.1f percent", 100*test_pop_count$n[map.gen["male"]]/sum(test_pop_count$n)));

# Total number of expected survivors
test_expected_surv = test_pop_count$n*surv_likelihood;
```

Total number of expected survivors in test set `r sum(test_expected_surv)`;

# Expected accuracy per gender

```{r, echo=FALSE, message=TRUE, warning=FALSE}
accuracy <- ifelse(surv_likelihood>0.5, surv_likelihood, 1-surv_likelihood)
 
print(sprintf("Optimal gender only predicted score            %2.4f ",  sum(test_pop_count$n*accuracy)/sum(test_pop_count$n)))
print(sprintf("Actual leader board (LB) score on the test set %2.4f ", 0.76555)); 
print(sprintf("Train set relative overstimation on LB         %2.4f ", (sum(test_pop_count$n*accuracy)/sum(test_pop_count$n)/0.76555))); 
```

# Create submission dataframe and output to file

The actual score verified after submission returns 0.76555.

```{r, echo=TRUE, message=TRUE, warning=FALSE}
submit <- data.frame(PassengerId = test_index, Survived = comb$Prediction[test_index]);
write.csv(submit, file = paste0("gender_only.csv"), row.names = FALSE, quote=F)
```