---
title: "Titanic, A joy trip turned into a disaster"
author: "Mukund G Kallapur"
date: "22-Jan-2019"
output:
  html_document:
    number_sections: true
    toc: true
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, error = TRUE)
```

# Introduction

This is my first Kaggle kernel. This is one of the most easiest datasets which can be found on Kaggle. This version of my kernel primarily 
focuses on the EDA or the Exploratory Data Analysis part of the raw data.


```{r Setup, message=FALSE}
## Load the library and read the file. 
library(tidyverse)
library(DMwR)

titanic_vis <- read.csv("../input/train.csv")
```
So, now that the data is loaded, let us move forward. 

# What this Titanic dataset about?

We all know that the Titanic ship accident was one of the biggest tragic incident of that time. This dataset contains demographic information 
about the passengers who travelled on that ill fated ship. This dataset is tremendously small in size as compared to some other 
humongous datasets we can see on Kaggle. 

Hence many beginners like me pick this dataset to begin with EDA, building basic models such as Logistic regression, slowly get into Machine
Learning part, Artificial Neural Networks and finally end up building models with Ensemble approaches. 

I believe that even before we think of modelling, EDA is must. Hence I have spent enough time and energy to come up with some decent visuals
with an interpretation behind each one of them. 

Before jumping directly to visuals, let me tell you what the Data tells us !

# Data Speaks, Let us hear !

```{r}
str(titanic_vis)
dim(titanic_vis)
```
By looking at the structure and the dmension, we can say that there are 891 records in the dataset with 11 Independent Variables and 
1 Target attribute which is the 'Survived' attribute represented as '1' and '0'. 

# Data Type conversion and recoding.

Notice that few of the attributes are considered as Integers by default by R. We have to convert those to categorical despite they being 
numbers, they represent categories. They will need recoding since it would give a better meaning when visualized.  


```{r}
### Conversion of the attributes 'Survived' and 'PClass' to 'Categorical'

titanic_vis$Survived <- as.factor(titanic_vis$Survived)
titanic_vis$Pclass <- as.factor(titanic_vis$Pclass)

### Recode the attributes Survived, PClass, Embarked and Gender for better Visualization. 
titanic_vis$Survived <-  ifelse(titanic_vis$Survived == "1","Yes", "No")
titanic_vis$Pclass <- ifelse(titanic_vis$Pclass == "1", "First",
                          ifelse(titanic_vis$Pclass == "2", "Second", "Third"))
titanic_vis$Sex <- ifelse(titanic_vis$Sex == "male", "Male", "Female")

titanic_vis$Embarked <- ifelse(titanic_vis$Embarked == "C", "Cherbourg",
                             ifelse(titanic_vis$Embarked == "Q", "Queenstown", 
                             ifelse(titanic_vis$Embarked == "S","Southampton","Southampton")))
                             
## Impute the Age

#summary(titanic_vis)
#sum(is.na(titanic_vis))
titanic_vis <- DMwR :: centralImputation(titanic_vis)
                             
## Binning the Age manually and splitting into groups using recode method. 

titanic_vis$Age <- ifelse(titanic_vis$Age <=1, "Baby",
                      ifelse((titanic_vis$Age > 1 & titanic_vis$Age <= 12), "Kid",
                             ifelse((titanic_vis$Age > 12 & titanic_vis$Age <= 19), "Teen",
                                    ifelse((titanic_vis$Age > 19 & titanic_vis$Age <= 40), "Young",
                                           ifelse((titanic_vis$Age > 40 & titanic_vis$Age <= 60), 
                                                  "Middle","Senior")))))
                                                  
titanic_vis$Age <- as.factor(titanic_vis$Age)                             
                             
```

# Visualizations 
## Overall Survival Count

The below plot highlights that out of 891 passengers, 342 were able to see the light of the day whereas 549 succumbed to death!

```{r}
titanic_vis %>%
  group_by(Survived) %>% 
      summarise(n=n()) %>% 
        ggplot() +
          geom_col(aes(reorder(Survived, -n),y=n, fill = Survived), width = 0.4) +
            xlab("Survival of Passengers") + ylab("Count") + 
              ggtitle("Titanic Survival Count") +
                  theme_minimal() + 
                    theme(plot.title = element_text(hjust = 0.5, face = "bold"),axis.title.x.bottom = element_text(face = "bold"), 
                        axis.title.y.left = element_text(face = "bold")) +
                            geom_text(aes(x = Survived, y = n, label = n), vjust = -0.5, stat = "sum", show.legend = FALSE)
```





## Survival Count against the Passenger Class 

Looking at the below plot we can infer that the passengers who were travelling in the least expensive Class 3 were the most affected. Out of 
491 passengers who travelled in 3rd class, only 119 were able to swim out alive !

```{r}
titanic_vis %>%
  group_by(Pclass, Survived) %>%
    dplyr :: summarise(n = n()) %>% 
      ggplot(aes(x = reorder(Pclass, -n), y = n, fill = Survived)) + 
          geom_bar(position = "dodge",stat = "identity", color = "black")  +
                geom_text(aes(label = n),position=position_dodge(width=0.9), vjust=-0.25) +
                    xlab("Passenger Class") + ylab("Count") + 
                        ggtitle("Titanic Passenger Class vs Survival") +
                            theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),axis.title.x.bottom = element_text(face = "bold",
  vjust = 3,size = 15), axis.title.y.left = element_text(face = "bold", size = 15),
  axis.text.x.bottom = element_text(face = "bold",size = 12,vjust = 3))
```

## Survival Count against Gender

Wow !! Indeed, many men lost their lives !!! 

```{r}
titanic_vis %>%
  group_by(Sex, Survived) %>%
    summarise(n = n()) %>% 
      ggplot(aes(x = reorder(Sex,-n), y = n, fill = Survived)) + 
          geom_bar(position = position_stack(), stat = "identity", width = 0.7, color = "black")  +
            geom_text(aes(label = n),position=position_stack(vjust = 0.5), size = 5) +
                             xlab("Passenger Gender") + ylab("Count") + 
                    ggtitle("Titanic Passenger Gender vs Survival") +
                      theme_minimal() +
  scale_fill_brewer(palette = 15) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),axis.title.x.bottom = element_text(face = "bold",vjust = 3,size = 15), axis.title.y.left = element_text(face = "bold", size = 15),axis.text.x.bottom = element_text(face = "bold",size = 12,vjust = 3))

```

## Survival Count against Embarked

Southampton being one of the major cities in the United Kingdom and also it being a major port there is no wonder that maximum number of
people embarked from there. Out of 891 passengers, a total of 646 people boarded the RMS Titanic from Southampton. Among those 646 passengers, 
427 survived to tell the tale !!
```{r}
titanic_vis %>%
  group_by(Embarked, Survived) %>%
  
  summarise(n = n()) %>%
  
  ggplot(aes(x = reorder(Embarked, -n), y = n, fill = Survived)) + 
  geom_bar(position = position_stack(), stat = "identity", width = 0.7, color = "black")  +

  geom_text(aes(label = n),position=position_stack(vjust = 0.5), size = 5) +
  xlab("Embarkation Point") + ylab("Count") + 
  ggtitle("Embarked against Survival") +
  theme_minimal() +
  scale_fill_brewer(palette = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.title.x.bottom = element_text(face = "bold",vjust = 3,size = 15), 
        axis.title.y.left = element_text(face = "bold", size = 15),
        axis.text.x.bottom = element_text(face = "bold",size = 12,vjust = 3))
```


## Survival Count against different Age Categories

Visualization of the below plot implies that people in the Age group belonging to 20 and 40, who are classified as "Young". The total population 
of people in this age range was 577. Among those who Survived were 208 and 369 died Young !

Another inference what we can draw from this plot is regariding the Babies less than 1 year. There were 14 babies and 12 survived. Kids in the 
age group 1 and 12 wre 55 in number and Survival ratio is almost equally distributed. 

Another staggering thing to notice is that there were 22 passengers aged 60 plus and among them only 5 survived. This may be due to the reason 
that they might have panicked when all the hell broke loose !!!
```{r}
titanic_vis %>%
  select(Age,Survived) %>%
    group_by(Age, Survived) %>%
        dplyr :: summarise(n = n()) %>%
            arrange(desc(Age)) %>%
  ggplot() +
  #coord_flip()+
        geom_col(aes(x=reorder(Age, -n), y = n,fill = Age)) +
          xlab("Age") + ylab("Count") + 
            facet_grid(~ fct_rev(Survived),scales = "free") +
                            guides(fill = guide_legend(title = "Age Categories"))  +
                    ggtitle("Survival vs Age")  +
                        theme_minimal() + 
                            theme(plot.title = element_text(hjust = 0.5, face = "bold"),
                                axis.title.x.bottom = element_text(face = "bold"), 
                                    axis.title.y.left = element_text(face = "bold"), 
                                        axis.text.x.bottom = element_text(face = "bold"),
                                            legend.title = element_text(face = "bold")) +
                                                geom_text(aes(x = Age, y = n, label = n, vjust = -0.5))
```

# Future Scope

In the future, I will be focussing on the Modelling part and build multiple models for Survival prediction.  

