---
title: "Titanic Passenger Survival"
author: "Daniel Vargas"
date: June 01, 2017
output: 
    html_document:
        theme: darkly
        code_folding: hide
---

## The Challenge

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

In this challenge, I will complete the analysis of what sorts of people were likely to survive.

## The data

The data has been split into two groups:

* training set (`train.csv`)
* test set (`test.csv`)

#### Data Dictionary

Variable	Definition	Key
survival	Survival	0 = No, 1 = Yes
pclass	Ticket class	1 = 1st, 2 = 2nd, 3 = 3rd
sex	Sex	
Age	Age in years	
sibsp	# of siblings / spouses aboard the Titanic	
parch	# of parents / children aboard the Titanic	
ticket	Ticket number	
fare	Passenger fare	
cabin	Cabin number	
embarked	Port of Embarkation	C = Cherbourg, Q = Queenstown, S = Southampton

### Understanding the data

```{r sets}
train <- read.csv("../input/train.csv", stringsAsFactors = FALSE)
test <- read.csv("../input/test.csv", stringsAsFactors = FALSE)
dim(train); dim(test)
str(train); str(test)
```

#### Quick Feature engineering

Splitting the cabins into desks to use as input

```{r feng}
strsplit(train$Cabin[2], NULL)[[1]]
train$Deck <- factor(sapply(train$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
test$Deck <- factor(sapply(test$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
train$Survived <- as.factor(train$Survived)
```

## Prediction model with Random Forests

```{r pilot, message = FALSE, warnings = FALSE}
set.seed(20170601)
library(e1071)
m <- naiveBayes(Survived ~ 
        Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Deck, 
        data = train, method = "rf", metric = "Accuracy", na.action = na.pass)
m
```
## Prediction and Solution

```{r prediction, message = FALSE, warnings = FALSE}
library(caret)
p <- predict(m, test, type = "class")
head(p, 20)
sol <- data.frame(PassengerID = test$PassengerId, Survived = p)
write.csv(sol, file = "titanic_sol.csv", row.names = F)
```