
---
title: 'Hands on case study in simple steps using R'
author: 'Supriya Rani'
date: '22 July 2018'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---


# Introduction
This is my first case study on kaggle. After spending hours for almost a month I am presenting my work here. My main purpose behind this script is to guide and inspire new kgglers.

## Understanding dataset
Titanic dataset is a tiny dataset for a beginner to start with. Here training dataset is provided to train a model and predict survivals on test dataset. General subbmission dataset is given as sample submission file. So let's start.

Main focus will be on the following topics:

* Feature Engineering (Applied machine learning)
* Data preparation
* Data Visualisation
* Modeling and Prediction

loading R packages as per need
```{r, message = FALSE}
# Loading library packages needed
library('randomForest') # classification algorithm
```

## Importing data files
Import dataset and assign it to new dataset.

```{r, message = FALSE, warning=FALSE}
# Import
titanic.train <- read.csv(file = "../input/train.csv", header = T, stringsAsFactors = F)
titanic.test <- read.csv(file = "../input/test.csv", header = T, stringsAsFactors = F)
```

## Exploring variables

```{r, message=FALSE, warning=FALSE}
# View data in different ways
str(titanic.train)
head(titanic.train)
tail(titanic.train)

str(titanic.test)
head(titanic.test)
tail(titanic.test)
```
We have details of datasets from which we can conclude the following:

Variable Name | Description
--------------|-------------
Passenger Id  | Unique identification Number of every passerger
Pclass        | Class of Passenger's traveling in 
Name          | Name of Passenger's
Sex           | Passenger's Sex
Age           | Age of Passenger's
SibSp         | Number of siblings/spouses present on that ship
Parch         | Number of parents/children present on that ship
Ticket        | Ticket number 
Fare          | Fare
Cabin         | Cabin Number
Embarked      | Port of embarkation
Survived      | For Survived (1) or for died (0)

We can see that titanic.train data have 891 observation and 12 variables. Where as titanic.test 418 observations and 11 variables. Survival variable is not present in titanic.test dataset. Also records are present in serial numbers divided in training and test dataset.

# Data preparation

## combining two files
To do feature engineering on both datasets we will combine them in a new dataset.
Since 'Survive' variable is missing in titanic.test dataset so we will add this variable with na values.

```{r, message=FALSE, warning=FALSE}
# creating variable 
titanic.test$Survived <- NA

# binding data row wise
titanic.combined <- rbind(titanic.train, titanic.test)

# check whether data ia bined correctly or not
head(titanic.combined)
tail(titanic.combined)
```
So 'titanic.combined' is new dataset with all the rows and top 891 rows are from titanic.train dataset and the bottom rest are from titanic.test dataset.

## Feature Engineering
### Data insights
This is very likely to happen that families might have survived or sinked together. So lets create a family variable to store family size.

```{r, message=FALSE, warning=FALSE}
# Create a new variable which contains siblings, spouse and individuals
titanic.combined$Family.size <- titanic.combined$SibSp + titanic.combined$Parch + 1
```

### Handeling missing values

```{r, message=FALSE, warning=FALSE}
# Check missing values
sapply(titanic.combined, function(x) {sum(is.na(x))})
```

-> Embarked variable

As only two value are missing so they are not going to effect our prediction model. Hence as 'S' has comparatively high presence so we can assign 'S' to those missing place.

```{r, message=FALSE, warning=FALSE}
# fill blan rows with 'S'

titanic.combined[titanic.combined$Embarked == '', "Embarked"] <- 'S' 
```
-> Age variable

We will apply simple linear regression model.

```{r, message=FALSE, warning=FALSE}
# modeling for age
age.formula <- "Age ~ Pclass + Sex"
age.model <- lm(
  formula = age.formula,
  data = titanic.combined
)
age.row <- titanic.combined[is.na(titanic.combined$Age), c("Pclass", "Sex")]

# predict age for NA filled rows
age.predict <-predict(age.model, newdata = age.row)

# assign value of age into combined dataset
titanic.combined[is.na(titanic.combined$Age), "Age"] <- age.predict
```
-> Fare variable
Since only one row has missing value so we will fill it with median

```{r, message=FALSE, warning=FALSE}
# find median of fare
fare.median <- median(titanic.combined$Fare, na.rm = T)

# assign value to blank rows
titanic.combined[is.na(titanic.combined$Fare), "Fare"] <- fare.median
```
## factoring character variables
So that we can use them later

```{r, message=FALSE, warning=FALSE}
titanic.combined$Pclass <- as.factor(titanic.combined$Pclass)
titanic.combined$Sex <- as.factor(titanic.combined$Sex)
titanic.combined$Embarked <- as.factor(titanic.combined$Embarked)
```

## Separate train and test data set
Now we have to split data so that we can use train dataset to *train* our model and to predict survivals for *test* dataset.
```{r, message=FALSE, warning=FALSE}
# split dataset
titanic.train <- titanic.combined[1:891,]
titanic.test <- titanic.combined[892:1309,]
```

# Modeling and Prediction  
## Preparing model
Now we are ready to train our model. Just to keep this script simple we will not include *Name* variable in our model, As here Name is a character variable so we have to apply feature enginering to break it in different buckets.

```{r, message=FALSE, warning=FALSE}
# To get Survived data in levels
titanic.train$Survived <- as.factor(titanic.train$Survived)

# Set a random seed
set.seed(675)

# Modeling data 
survived.formula <- as.formula("Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Family.size")
survive.model <- randomForest(formula = survived.formula, data = titanic.train, ntree = 500, mtry = 3, nodesize = 0.01 * nrow(titanic.test),  keep.forest = TRUE)
```

## Prediction
For prediction of survivals we will proceed with *Test* dataset

```{r, message=FALSE, warning=FALSE}
# Prediction 
Survived <- predict(survive.model, newdata = titanic.test)
```

## Creating new data set for submission
As per given instructions we have to submit file in csv format which will only include *PassengerId* and *Survival* from *Test* dataset.
```{r, message=FALSE, warning=FALSE}
# new data set with two variables
PassengerId <- titanic.test$PassengerId
output.df <- as.data.frame(PassengerId)
output.df$Survived <- Survived
table(output.df$Survived)
# write csv file

write.csv(output.df, file = "titanic_submission.csv", row.names = FALSE)
```

# Visualisation of final dataset 
```{r, message=FALSE, warning=FALSE}
# error visuals
plot(survive.model, ylim=c(0,0.40))
legend('topright', colnames(survive.model$err.rate), col=1:3, fill=1:3)
  ```
  
  Here Green and red lines shows survived and died respectively. Black line shows error which is near to 20%. Hence our model is good enough to go. 
  
# Conclusion
This is my first exploration of a Kaggle dataset. I look forward to doing more. Comments and feedbacks are welcome!












