
## Title: "Titanic Survival Prediction via RF"
#### Author: Rahima Karimova (Inspired by David Langer's Titanic 
#### Data Analysis video https://www.youtube.com/watch?v=32o0DnuRjfg 
#### and kaggler-Megan Risdal)
#### Date: 29.11.2016

### Necessary packages 
#### library(tidyverse) 
#### library(ggthemes)
#### library(stringr)
#### library(scales)
#### library(mice)
#### library(randomForest)
```{r include=FALSE}
library(tidyverse) #Cleaning, visualization...
library(ggthemes) # visualisation
library(stringr)
library(scales)
library(mice) # NA replacement
library(randomForest) # forecast
```
# Reading your data
```{r}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)
```

# Discovering your data
```{r}
glimpse(train)
glimpse(test)
```
## Preparing train and test for binding
```{r}
test.survived<-data.frame(Survived= rep("None", nrow(test)), test[,])
head(test.survived)
ntest.survived<-test.survived[,c(2,1,3,4,5,6,7,8,9,10,11,12)]
head(ntest.survived)
names(train)
names(ntest.survived)
```

## Binding data sets
```{r}
data.combined<-rbind(train,ntest.survived)
glimpse(data.combined)

data.combined$Survived<-as.factor(data.combined$Survived)
data.combined$Pclass<-as.factor(data.combined$Pclass)
glimpse(data.combined)
```
## Visualisation part 1

### Survival distribution
```{r}
table(data.combined$Survived)
```
### Distribution around classes
```{r}
table(data.combined$Pclass)
```

### Proportion of SEX by PClass
```{r,echo=T}
g2.3<-ggplot(data.combined,aes(x=factor(Pclass),fill=factor(Sex)))+geom_bar(position="fill")
g2.3
```
### Rich folks survival rate higher? Looking at Pclass and Sex
```{r,echo=T}
g1<-ggplot(test,aes(x=(Pclass),fill=(Sex)))+geom_bar(width=0.5)
g1
g2<-ggplot(train,aes(x=factor(Pclass),fill=factor(Sex)))+geom_bar(position="dodge")+facet_grid(.~Survived)
g2
```
## Survival rate depending on Age, Pclass and Sex
```{r}
posn.j <- position_jitter(0.5, 0)
g3<-ggplot(train,aes(x=factor(Pclass),y=Age,col=factor(Sex)))+geom_jitter(position=posn.j,alpha=0.5,size=3)+facet_grid(.~Survived)
g3
```

#Data Cleaning

## How many unique names are there in both test and train datasets?
```{r}
length(unique(data.combined$Name))
str(data.combined)
```
### So we have 1307 unique names vs. 1309 names. Lets see if there are duplicates.
```{r}
dup.names<-as.character(data.combined[which(duplicated(as.character(data.combined$Name))),"Name"])
data.combined[which(data.combined$Name %in% dup.names),]
```
# Feature extraction

### Any valuable information in the Title (Mrs, Miss, Mr, Master)?

```{r}
misses<-data.combined[which(str_detect(data.combined$Name,"Miss.")),]
head(misses)
mrses<-data.combined[which(str_detect(data.combined$Name,"Mrs.")),]
mrses[1:5,]
masters<-data.combined[which(str_detect(data.combined$Name,"Master")),]
masters[1:5,]
mr<-data.combined[which(str_detect(data.combined$Name,"Mr.")),]
mr[1:5,]
```
## Adding new variable - Title
```{r}
extractTitle<-function(Name){
  Name<-as.character(Name)
  if(length(grep("Miss.",Name))>0){
    return("Miss.")
  } else if (length(grep("Master.",Name))>0){
    return("Master.")
  } else if(length(grep("Mrs.",Name))>0){
    return("Mrs.")
  } else if(length(grep("Mr.", Name))>0){
    return("Mr.")
  } else {
    return("Other")
  }
}
titles<-NULL
for (i in 1:nrow(data.combined)){
  titles<-c(titles,extractTitle(data.combined[i,"Name"]))
}
```
### Ok. Lets factorise our new feature
```{r}
data.combined$Title<-as.factor(titles)
```
# Visualisation part 2
### Lets see how Passenger class and Title is influencing the survival rate
```{r, echo=T}
p2.1<-ggplot(data.combined[1:891,],aes(x=Title,fill=Survived))+geom_bar(width=0.5)+
  facet_wrap(~Pclass)+
  ggtitle("Pclass")+
  xlab("Title")+
  ylab("Total Count")+
labs(fill="Survived")

p2.1
```
## Whats the distribution of females and males in test and train sets?
```{r}
table(data.combined$Sex)
```
```{r}
p2.2<-ggplot(data.combined[1:891,],aes(x=Sex,fill=Survived))+geom_bar(width=0.5)+
  facet_wrap(~Pclass)+
  ggtitle("Pclass")+
  xlab("Title")+
  ylab("Total Count")+
  labs(fill="Survived")

p2.2

summary(data.combined$Age)
```
## Is there any pattern among Male and Female passengers and Pclass?
```{r, echo=T}
g2.3<-ggplot(data.combined[1:891,],aes(x=Age,fill=Survived))+geom_bar(width=0.5)+
  facet_wrap(~Sex + Pclass)+
  geom_bar(binwidth = 10)+
  xlab("Age")+
  ylab("Total Count")

g2.3
```

### Validate that Master is a good proxy for male children 
```{r}
boys<-data.combined[which(data.combined$Title=="Master."),]
summary(boys$Age)
```
### Miss is more complicated,lets look
```{r}
misses<-data.combined[which(data.combined$Title=="Miss."),]
summary(misses$Age)
```
### Is there any pattern within Misses between Pclass?
```{r, echo=T}
g2.4<-ggplot(misses[misses$Survived !="None",],aes(x=Age,fill=Survived))+
  facet_wrap(~Pclass)+
  geom_bar(binwidth=5)+
  ggtitle("Age for Miss by Pclass")+
  xlab("Age")+
  ylab("Total Count")

g2.4
```
### Appears female children have diffrent survival rate
```{r}
misses.alone<-misses[which(misses$SibSp==0, misses$Parch==0),]
summary(misses.alone$Age)
```
### move on Subsp variable
```{r}
summary(data.combined$SibSp)
```
### Factoring Sibsp variable to further continue with histogram plotting
```{r}
data.combined$SibSp<-as.factor(data.combined$SibSp)
```
## Visualization of the survival rate by Sibsp, Title and Pclass
```{r,echo=T}
g2.5<-ggplot(data.combined[1:891,],aes(x=SibSp, fill=Survived))+geom_bar(width=0.5)+
facet_wrap(~Pclass+Title)+
  ggtitle("Pclass,Title")+
  xlab("Sibsp")+
  ylab("Total Count")+
  ylim(0,300)+
  labs(fill="Survived")

g2.5
```

### Factoring Parch variable
```{r}
data.combined$Parch<-as.factor(data.combined$Parch)
```
## Visualization of the Survival rate by Parch, Title and Pclass
``` {r,echo=T}
g2.6<-ggplot(data.combined[1:891,],aes(x=Parch, fill=Survived))+geom_bar(width=1)+
   facet_wrap(~Pclass+Title)+
   ggtitle("Pclass,Title")+
   xlab("Parch")+
   ylab("Total Count")+
   ylim(0,300)+
   labs(fill="Survived")
  
g2.6
``` 
## Creating the Family Size feature
```{r}
  temp.sibsp<-c(train$SibSp,test$SibSp)
  temp.parch<-c(train$Parch,test$Parch)
  data.combined$Family.Size<-as.factor(temp.sibsp+temp.parch+1)
```
```{r,echo=T}
g2.7<-ggplot(data.combined[1:891,],aes(x=Family.Size, fill=Survived))+geom_bar(width=1)+
    facet_wrap(~Pclass+Title)+
    ggtitle("Pclass,Title")+
    xlab("Family.Size")+
    ylab("Total Count")+
    ylim(0,300)+
    labs(fill="Survived")
  
  g2.7
```
# Working on NAs.
##Constructing subgroups
```{r}
subsmr<-subset(data.combined[1:891,], is.na(Age) & Title=="Mr.")
summary(subsmr)
```
```{r,echo=T}
g3.1<-ggplot(subsmr,aes(x=Pclass,fill=Survived))+geom_bar(width=0.5)+facet_wrap(~Family.Size)
g3.1
```   
```{r}  
summary(misses.alone$Age)
```
### Here we see the age range is wide so we need futher classification
  
  
## Substituting NAs with means
```{r} 
   data.combined2<-data.combined
   
   data.combined2$Embarked[c(62,830)]
   data.combined2$Embarked[c(62,830)]<-"C"
   data.combined2[1044,]
```   
### Replacement of missing fare value with median fare for class/embarkment
```{r}
   data.combined2$Fare[1044]<-median(data.combined2[data.combined2$Pclass==3 & data.combined2$Embarked=="S",]$Fare,na.rm=T)
   #Make variables factors into factors
   factor_vars <- c('PassengerId','Pclass','Sex','Embarked',
                    'Title','Family.Size')
   
   data.combined2[factor_vars] <- lapply(data.combined2[factor_vars], function(x) as.factor(x))
```
```{r}
   ### Setting a random seed
   set.seed(129)
```
### Performing mice imputation, excluding certain less-than-useful variables:
```{r}
mice_mod <- mice(data.combined2[, !names(data.combined2) %in% c('PassengerId','Name','Ticket','Cabin','Survived')], method='rf') 
### complete output saving
```{r}
   mice_output <- complete(mice_mod)
```  
   ## Ploting age distributions
```{r}
par(mfrow=c(1,2))
   hist(data.combined2$Age, freq=F, main='Age: Original Data', 
        col='darkgreen', ylim=c(0,0.04))
   hist(mice_output$Age, freq=F, main='Age: MICE Output', 
        col='lightgreen', ylim=c(0,0.04))
```
   
### Replacement of Age variable from the mice model.
```{r}
data.combined2$Age <- mice_output$Age
```

### Replacement of Deck variable from the mice model.
  ```{r}
data.combined2$Deck <- mice_output$Deck
summary(data.combined2)
```
# Prediction

## Splitting the data back into a train set and a test set
```{r}
   train <- data.combined2[1:891,]
   test <- data.combined2[892:1309,]
```   
### Set a random seed
```{r}
   set.seed(754)
```
   
   ## Building the model (note: not all possible variables are used)
   ```{r}
   rf_model <- randomForest(factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + 
                              Fare + Embarked + Title,ntree=100, importance=T,
                               data = train)
   ```
   ## Showing model error
   ```{r,echo=T}
   plot(rf_model, ylim=c(0,0.36))
   legend('topright', colnames(rf_model$err.rate), col=1:3, fill=1:3)
  ```
  
   ## Getting importance
   ```{r}
   importance    <- importance(rf_model)
   varImportance <- data.frame(Variables = row.names(importance), 
                               Importance = round(importance[ ,'MeanDecreaseGini'],2))
  ```
   ## Creating a rank variable based on importance
   ```{r}
   rankImportance <- varImportance %>%
     mutate(Rank = paste0('#',dense_rank(desc(Importance))))
  ``` 
   ### Variables relative importance visualisation
   ```{r,echo=T}
   ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
                              y = Importance, fill = Importance)) +
     geom_bar(stat='identity') + 
     geom_text(aes(x = Variables, y = 0.5, label = Rank),
               hjust=0, vjust=0.55, size = 4, colour = 'red') +
     labs(x = 'Variables') +
     coord_flip() + 
     
     theme_few()
    ```
   #Prediction
   
   ##Prediction using the test set
   ```{r}
   prediction <- predict(rf_model, test)
   ```
   ## Saving the solution to a dataframe with two columns: PassengerId and Survived (prediction)
   ```{r}
   solution <- data.frame(PassengerID = test$PassengerId, Survived = prediction)
   ```
   ## Writing the solution to file
   ```{r}
   write.csv(solution, file = 'rf_mod_Solution.csv', row.names = F)
   ```
