---
title: 'Exploring the Titanic Dataset'
author: 'Megan L. Risdal'
date: '6 March 2016'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is my first stab at a Kaggle script. I have chosen to work with the Titanic dataset after spending some time poking around on the site and looking at other scripts made by other Kagglers for inspiration. I will also focus on doing some illustrative data visualizations along the way. I'll then use `randomForest` to create a model predicting survival on the Titanic. I am new to machine learning and hoping to learn a lot, so feedback is very welcome! 

There are three parts to my script as follows:

* Feature engineering
* Missing value imputation
* Prediction!

## Load and check data

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
library('tutorial')
```

```{r, include=FALSE}
tutorial::go_interactive()
```

Now that our packages are loaded, let's read in and take a peek at the data.

```{r, message=FALSE, warning=FALSE}
train <- read.csv('http://s3.amazonaws.com/assets.datacamp.com/course/importing_data_into_r/swimming_pools.csv', stringsAsFactors = F)
head(train)


# check data
str(full)
```
