---
title: "Titanic in Carrots"
author: "Srikant Jayaraman"
date: "5/9/2018"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)

```

## Load Data and Packages
# Load Packages ====
```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(stringr)
library(mice)
library(randomForest)
library(caretEnsemble)
library('ggthemes')
library('scales') 
```

## Load Data ====

```{r , echo=FALSE}
train_data <- read.csv("../input/train.csv", stringsAsFactors = F)
test_data <- read.csv("../input/test.csv", stringsAsFactors = F)

full_data <- bind_rows(train_data, test_data)
str(full_data)

```

## Transform Data ====
# Make some varibales factors ====
```{r}


```
# Introduce Title Variable
```{r}

full_data$Title <- (gsub('(.*, )|(\\..*)', '', full_data$Name))

# Show title counts by sex
table(full_data$Sex, full_data$Title)

# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
full_data$Title[full_data$Title == 'Mlle']        <- 'Miss' 
full_data$Title[full_data$Title == 'Ms']          <- 'Miss'
full_data$Title[full_data$Title == 'Mme']         <- 'Mrs' 
full_data$Title[full_data$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full_data$Sex, full_data$Title)

full_data$Title <- as.factor(full_data$Title)

```


# Finally, grab surname from passenger name
```{r}
full_data$Surname <- sapply(full_data$Name,  
                            function(x) strsplit(x, split = '[,.]')[[1]][1])

full_data$Surname <- as.factor(full_data$Surname)
str(full_data)

```


# Make Family Variables ====

# Family Size and then create a family variable
```{r}
full_data$family_size <- full_data$SibSp + full_data$Parch + 1

# Create a family variable 
full_data$Family <- as.factor(paste(full_data$Surname, full_data$family_size, sep='_'))
```


# Explore the family and family size variables ====

```{r}
ggplot(full_data[1:891,], aes(x = family_size, fill = factor(Survived))) +
  geom_bar(position='fill') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()

```

So, families of size greater than 7 did not survive at all. Let us discretize this thing.
# Discretize family size

```{r}

# full_data$family_size_group[full_data$family_size == 1] <- 'single'
# full_data$family_size_group[full_data$family_size %in% 2:4] <- 'two_to_four'
# full_data$family_size_group[full_data$family_size %in% 5:7] <- 'five_to_seven'
# full_data$family_size_group[full_data$family_size >= 8] <- 'gt_eight'


full_data$family_size_group[full_data$family_size == 1] <- 'singleton'
full_data$family_size_group[full_data$family_size < 5 & full_data$family_size > 1] <- 'small'
full_data$family_size_group[full_data$family_size > 4] <- 'large'
table(full_data$family_size_group)
# Factorize Family Size Group Variable 

full_data$family_size_group <- as.factor(full_data$family_size_group)
#levels(full_data$family_size_group) = c('single', 'two_to_four', 'five_to_seven', 'gt_eight')
levels(full_data$family_size_group) = c('single', 'medium', 'large')

# Now plot survival
full_data %>% filter(!is.na(Survived)) %>% ggplot(aes(x = family_size, fill = factor(Survived))) +
  geom_bar(position='fill') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()

mosaicplot(table(full_data$family_size_group, full_data$Survived), main='Family Size by Survival', shade=TRUE)


```

```{r}

full_data$Cabin_letter <- as.factor(str_extract(full_data$Cabin, boundary("character")))
full_data$Cabin_letter_length <- str_length(full_data$Cabin)
```


# Now treat Missing variables ====

THe Variables Deck, Age and 1 missing Value in Fare

# Replace Missing Fare
```{r}
summary(full_data)

full_data %>% filter(Pclass == 3) %>% ggplot(aes(x = Fare)) + geom_density()

full_data$Fare[is.na(full_data$Fare)] <- median(full_data$Fare[full_data$Pclass == 3 & full_data$Embarked == "S"], na.rm = TRUE)

```


# Replace missing embarkment ====
```{r}

embark_fare <- full_data %>%
  filter(PassengerId != 62 & PassengerId != 830)

# Use ggplot2 to visualize embarkment, passenger class, & median fare
ggplot(embark_fare, aes(x = Embarked, y = Fare, fill = factor(Pclass))) +
  geom_boxplot() +
  geom_hline(aes(yintercept=80), 
             colour='red', linetype='dashed', lwd=2) +
  scale_y_continuous(labels=dollar_format()) +
  theme_few()

full_data$Embarked[c(62, 830)] <- "C"

full_data$Sex <- as.factor(full_data$Sex)
full_data$Embarked <- as.factor(full_data$Embarked)
full_data$PassengerId <- as.factor(full_data$PassengerId)
full_data$Pclass <- as.factor(full_data$Pclass)

```

# Mice Imputation for Ages and Cabins ====
```{r}
# summary(full_data)
# sapply(full_data, class)

# Set a random seed
set.seed(129)

# Perform mice imputation, excluding certain less-than-useful variables:
mice_mod <- mice(full_data[, !names(full_data) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname','Survived', 'Cabin_letter_length')], method='rf')

mice_output <- complete(mice_mod)

# Repalce Age and Cabin_letter with imputed values
full_data$Age <- mice_output$Age
#full_data$Cabin_letter <- as.factor(mice_output$Cabin_letter)

# Look at Mice output
par(mfrow = c(1, 1))
hist(mice_output$Age)
hist(full_data$Age)
```

Now that looks good. Lets go further
# create Mother and Child variable because she does it!
```{r}

full_data$Child[full_data$Age < 18] <- 'Child'
full_data$Child[full_data$Age >= 18] <- 'Adult'

# Show counts
table(full_data$Child, full_data$Survived)

# Adding Mother variable
full_data$Mother <- 'Not Mother'
full_data$Mother[full_data$Sex == 'female' & full_data$Parch > 0 & full_data$Age > 18 & full_data$Title != 'Miss'] <- 'Mother'

# Show counts
table(full_data$Mother, full_data$Survived)

full_data$Child <- as.factor(full_data$Child)
full_data$Mother <- as.factor(full_data$Mother)

```

# Transform Fare ====
```{r}

full_data$Fare_bin <- (full_data$Fare - min(full_data$Fare))/(max(full_data$Fare)-min(full_data$Fare))
full_data$num_people_with_ticket <- sapply(full_data$Ticket, function(x){sum(full_data$Ticket == x)})

full_data$party_size[full_data$num_people_with_ticket == 1] <- "singleton"
full_data$party_size[full_data$num_people_with_ticket %in% 2:4] <- "two_to_four"
full_data$party_size[full_data$num_people_with_ticket %in% 5:8] <- "five_to_eight"
full_data$party_size[full_data$num_people_with_ticket > 8] <- "gt_eight"

full_data$party_size <- as.factor(full_data$party_size)


# full_data$num_survivors_in_ticket 
# 
# a <- sapply(full_data$Ticket[3], function(x){(sum(full_data$Survived[full_data$Ticket == x & full_data$PassengerId != full_data$PassengerId[full_data$Ticket]] == 1, na.rm = TRUE)) })

surviving_tickets <- unique(full_data$Ticket[full_data$Survived == 1 & !is.na(full_data$Survived) &full_data$party_size != "singleton"])

full_data$surviving_ticket <- sapply(full_data$Ticket, function(x){x %in% surviving_tickets})


# Show that people survive or die together 
full_data %>% 
  filter(!is.na(Survived)) %>% 
  ggplot(aes(x = surviving_ticket, 
             fill = factor(Survived))) + geom_bar(position = "fill")
```

# So, now for predictions ====

# Split the data back into a train set and a test set
```{r}

train <- full_data[1:891,]
test <- full_data[892:1309,]

```

# Build the Models now ===
```{r}
# Set a random seed
set.seed(754)

# Build the model (note: not all possible variables are used)
rf_model <- randomForest(factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + 
                           Fare_bin + party_size + surviving_ticket+Embarked + Title + family_size_group + Child + Mother,
                         data = train)


sapply(full_data, class)

# Show model error
plot(rf_model, ylim=c(0,0.36))
```


# Get importance
```{r}

importance    <- importance(rf_model)
varImportance <- data.frame(Variables = row.names(importance), 
                            Importance = round(importance[ ,'MeanDecreaseGini'],2))

# Create a rank variable based on importance
rankImportance <- varImportance %>%
  mutate(Rank = paste0('#',dense_rank(desc(Importance))))

# Use ggplot2 to visualize the relative importance of variables
ggplot(rankImportance, aes(x = reorder(Variables, Importance), 
                           y = Importance, fill = Importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
            hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'Variables') +
  coord_flip() + 
  theme_few()


```

# Make predictions ====
```{r}
p <- predict(rf_model, test)
results <- data.frame(PassengerId = test_data$PassengerId, Survived = p)

write.csv(results, file = "submit_results.csv", row.names = FALSE)
```


