
This is my very initial code, it may not be the best prediction but it will definitely 
help absolute beginners get started!

**Getting Started**

We load the library required followed by the dataset.

```{r}
library(caret) 

train<-read.csv("../input/train.csv", 
                na.strings=c("NA", "#DIV/0!"))
```

To keep this simple, we drop any columns which are less informative for the modelling and
impute missing values with the mean.

```{r}
training1<-subset(train, select = -c(PassengerId, Name, Ticket, Cabin))
missingcol<-data.frame(colMeans(is.na(training1)))
mean(training1$Age, na.rm=TRUE)
mean(training1$Fare,na.rm=TRUE)
training1$Age[is.na(training1$Age)] <-29.69 

```

*Remember* to set Categorical column as factor before modelling.
```{r}
training1$Sex<-as.factor(training1$Sex)
training1$Embarked<-as.factor(training1$Embarked)
training1$Survived<-as.factor(training1$Survived)
```

**Data Visualization**  

Investigate for any outliers or trends in the data using Scatterplots and Histogram.
No observations was observed.
```{r}
par(mfrow=c(2,2))
plot(training1$Pclass, training1$Fare)
hist(training1$Age)
hist(training1$SibSp)
hist(training1$Parch)
```

**Validation**  

Further split the data to 80% training and 20% validate dataset so that testing of the 
model predictions can be made before prediction is made to the test set.
```{r}
inBuild<-createDataPartition(y=training1$Survived, p=0.8, list=FALSE)
trainData<-training1[inBuild,]
validateData<-training1[-inBuild,]
```
**CART**  

First, we use Classification and Regression Tree for the modelling. We use the model fit
to predict the validate dataset to calculate the accuracy of the model. The CART model
give approx 76% accuracy.
```{r}
set.seed(1120345)
modelfit_rpart<-train(Survived~., data=trainData, method="rpart")
predict_rpart<-predict(modelfit_rpart, newdata=validateData)
result_rpart<-confusionMatrix(predict_rpart, validateData$Survived)
result_rpart
```

**Random Forest**  

Similarly, we use Random Forest which gives us an accuracy of approx 80%,
```{r}
modelfit_rf<-train(Survived~., data=trainData, method="rf", prox=TRUE, 
                    trControl=trainControl(method = "cv", number=5))

predict_rf<-predict(modelfit_rf, newdata=validateData)
result_rf<-confusionMatrix(predict_rf, validateData$Survived)
result_rf
```

**Generalized Boosted Regression Model**  

Next, we tried modelling with Generalized Boosted Regression Model which gives us 
accuracy of approx 82%
```{r}
modelfit_gbm<-train(Survived~., data=trainData, method="gbm", verbose=FALSE, trControl=trainControl(method="cv", number=5))
predict_gbm<-predict(modelfit_gbm, newdata=validateData)
result_gbm<-confusionMatrix(predict_gbm, validateData$Survived)
result_gbm
```

Finally, we decided to use Generalized Boosted Regression Model to predict the test 
dataset and submit the prediction.


