---
title: "Easy R Titanic Kernel (Please Upvote if you like or learned from my Kernel here)"
author: "Roger Rosales"
output:
  html_document:
    number_sections: true
    toc: true
---

I really appreciate if you <strong>UPVOTE</strong> this kernel and my other kernels. I am trying to get up to Kaggle Expert. Thank you!

This R markdown predictive model for the Titanic dataset is more about the process of creating a Kaggle Kernel, 
how to read in data using Kaggle's pathways, and how to export your predictions for submission.
I hope you enjoy it!


# Load in the Titanic datasets

```{r}
titanic.train <- read.csv("../input/train.csv");
titanic.test  <- read.csv("../input/test.csv");
```

## Check that the datasets looks okay.
```{r}
head(titanic.train)
tail(titanic.train)
summary(titanic.train)
str(titanic.train)


head(titanic.test)
tail(titanic.test)
summary(titanic.test)
str(titanic.test) #no survived variable in the test set
```


# Time to clean up the data

1. Combine the two datasets and cleanup as one dataset

2. Look for missing values

3. Change the types of the columns if needed




## Let's combine the two files and then do the cleanup.

We will create a flag column to indicate which row is coming from training vs. the test set.

```{r}
titanic.train$IsTrainSet <- TRUE
titanic.test$IsTrainSet <- FALSE
```

Before we can combine the two datasets, they have to have the same columns.

But the test set is missing Survived.

We can create a Survived column in the test set.

First check that the column names are the same expect for Survived

```{r}
names(titanic.train)
names(titanic.test)
```

Yes the column names match.

Now we can create the Survived column in the test set.

```{r}
titanic.test$Survived <- NA
```

Now we can combine the two datasets by rbind.

```{r}
titanic.full <- rbind(titanic.train , titanic.test)
```

Let's make a table() on IsTrainSet

There should be 891 in the training and 418 in the test.

```{r}
table(titanic.full$IsTrainSet)
```


## Now let's look for missing values.

Investigate Embarked column

```{r}
table(titanic.full$Embarked)
```

There are 2 missing values. Also note that the mode is "S".

Filter your data to Embarked = '' and replace with the mode of S.

```{r}
titanic.full[titanic.full$Embarked == "",] 
```

Both first class females are from Cabin B28.

```{r}
table(titanic.full$Sex,titanic.full$Embarked)
table(titanic.full$Pclass,titanic.full$Embarked)
```

Consideration: Build a model to predict their Embarked.

```{r}
titanic.full[titanic.full$Embarked == "","Embarked"] <- "S"
table(titanic.full$Embarked)
```

No more missing values!

Investigate the Age column.

```{r}
table(is.na(titanic.full$Age))
```

There are 263 values are missing age. Not good!

Replace the missing Age values with median.

Run median function, make sure to remove NAs in the calculation.

```{r}
median(titanic.full$Age,na.rm = TRUE)
```

The median is 28.

Try assigning ages based on gender and adult vs. child.

If you are 17+ and male, you are called Mr.

If you are a female and Miss title, you are not married.

```{r}
titanic.males <- titanic.full[titanic.full$Sex == "male",]

median(titanic.males$Age,na.rm = TRUE)

medianAgeMale <- median(titanic.males$Age,na.rm = TRUE)

titanic.full[is.na(titanic.full$Age)&titanic.full$Sex == 'male',
             "Age"] <- medianAgeMale

titanic.females <- titanic.full[titanic.full$Sex == "female",]

median(titanic.females$Age,na.rm = TRUE)

medianAgeFemale <- median(titanic.females$Age,na.rm = TRUE)

titanic.full[is.na(titanic.full$Age)&titanic.full$Sex == 'female',
             "Age"] <- medianAgeFemale
```


Check for missing values again.
```{r}
table(is.na(titanic.full$Age))
```

Investigate the Fare column.

```{r}
table(is.na(titanic.full$Fare))
```

One missing FARE value.


Let's create a decision tree to determine the missing fare.

```{r}
library("rpart")

library("rpart.plot")

fare_dt <- rpart(Fare ~ Pclass + Parch + Sex + SibSp,
                 data=titanic.full)

rpart.plot(fare_dt, type=1, extra=101)

MissingFare <- titanic.full[is.na(titanic.full$Fare),]
MissingFare

titanic.full[is.na(titanic.full$Fare),"Fare"] <- 12
```

Check the missing FARE again.

```{r}
table(is.na(titanic.full$Fare))
```


## Change the types of the columns if needed.

Change Pclass, Sex, and Embarked to factor.

```{r}
str(titanic.full)
titanic.full$Pclass <- as.factor(titanic.full$Pclass)
titanic.full$Sex <- as.factor(titanic.full$Sex)
titanic.full$Embarked <- as.factor(titanic.full$Embarked)
```




# Split the full data out again

```{r}
titanic.train <- titanic.full[titanic.full$IsTrainSet == TRUE,]
titanic.test <- titanic.full[titanic.full$IsTrainSet == FALSE,]
nrow(titanic.train)
nrow(titanic.test)
```

Set the Survived column in the train set.

```{r}
titanic.train$Survived <- as.factor(titanic.train$Survived)
str(titanic.train)
```

# Let's visualize some data
```{r}
library(ggplot2)
g <- ggplot(data=titanic.train, aes(x=Sex,fill=Survived)) + geom_bar()+
  theme_grey(base_size = 14)
g

g <- ggplot(data=titanic.train, aes(x=Pclass,fill=Survived)) + geom_bar()+
  theme_grey(base_size = 14)
g

g <- ggplot(data=titanic.train, aes(x=Embarked,fill=Survived)) + geom_bar() +
  theme_grey(base_size = 14)
g

g <- ggplot(data=titanic.train, aes(x=SibSp,fill=Survived)) + geom_bar() +
  theme_grey(base_size = 14)
g

g <- ggplot(data=titanic.train, aes(x=Age)) + geom_density()
g

g <- ggplot(data=titanic.train, aes(x=Fare)) + geom_density()+
  facet_grid(~Pclass)
g

g <- ggplot(data=titanic.train, aes(x=Pclass,y=Fare))+
  geom_boxplot()

g

# Jitter Plot by Class and Survived
g <- ggplot(data=titanic.train, aes(x=Pclass,y=Fare,
                                    color=Survived))+ geom_jitter() + ggtitle("Titanic Survival by Class")

g


# Jitter Plot by Class, Gender and Survived
#Same chart but for Males only
g <- ggplot(data=titanic.train[titanic.train$Sex == "male",], aes(x=Pclass,y=Fare,
                                    color=Survived))+ geom_jitter() + ggtitle("Titanic Survival by Class: Males Only")

g


#Same chart but for Females only
g <- ggplot(data=titanic.train[titanic.train$Sex == "female",], aes(x=Pclass,y=Fare,
                                    color=Survived))+ geom_jitter() + ggtitle("Titanic Survival by Class: Females Only")

g




```

Wow look at those Females who were in the 3rd class. Many of them didn't survive!


# Create the random forest predictive model


```{r message=FALSE, warnings=FALSE}
library("randomForest")
```

```{r}
survived.equation <- "Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked"
survived.formula <- as.formula(survived.equation)


titanic.model <- randomForest(formula=survived.formula,
             data=titanic.train,
             ntree = 500,
             mtry = 3,
             nodesize = 0.01*nrow(titanic.train)
             )

Survived <- predict(titanic.model,newdata=titanic.test)
```

# Get ready to submit predictions into Kaggle

```{r}
PassengerId <- titanic.test$PassengerId
output.df <- as.data.frame(PassengerId)

output.df$Survived <- Survived

write.csv(output.df,"titanic_kaggle_submission.csv",row.names = FALSE)
```

Again, I really appreciate if you <strong>UPVOTE</strong> this kernel and my other kernels. I am trying to get up to Kaggle Expert. Thank you!