---
title: 'Exploring the Titanic Dataset'
author: 'Megan L. Risdal'
date: '6 March 2016'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is my first stab at a Kaggle script. I have chosen to work with the Titanic dataset after spending some time poking around on the site and looking at other scripts made by other Kagglers for inspiration. I will also focus on doing some illustrative data visualizations along the way. I'll then use `randomForest` to create a model predicting survival on the Titanic. I am new to machine learning and hoping to learn a lot, so feedback is very welcome! 

There are three parts to my script as follows:

* Feature engineering
* Missing value imputation
* Prediction!

## Load and check data

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
```

Now that our packages are loaded, let's read in and take a peek at the data.

```{r, message=FALSE, warning=FALSE}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)

full  <- bind_rows(train, test) # bind training & test data

# check data
str(full)
```

We've got a sense of our variables, their class type, and the first few observations of each. We know we're working with 1309 observations of 12 variables. To make things a bit more explicit since a couple of the variable names aren't 100% illuminating, here's what we've got to deal with:

Variable Name | Description
--------------|-------------
Survived      | Survived (1) or died (0)
Pclass        | Passenger's class
Name          | Passenger's name
Sex           | Passenger's sex
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation


# Feature Engineering
## What's in a name?

The first variable which catches my attention is **passenger name** because we can break it down into additional meaningful variables which can feed predictions or be used in the creation of additional new variables. For instance, **passenger title** is contained within the passenger name variable and we can use **surname** to represent families. Let's do some **feature engineering**!

```{r, message=FALSE, warning=FALSE}
# Grab title from passenger names
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Show title counts by sex
table(full$Sex, full$Title)

# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
full$Title[full$Title == 'Mlle']        <- 'Miss' 
full$Title[full$Title == 'Ms']          <- 'Miss'
full$Title[full$Title == 'Mme']         <- 'Mrs' 
full$Title[full$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full$Sex, full$Title)

# Finally, grab surname from passenger name
full$Surname <- sapply(full$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])
```

```{r results='asis'}
cat(paste('We have <b>', nlevels(factor(full$Surname)), '</b> unique surnames. I would be interested to infer ethnicity based on surname --- another time.'))
```

## Do families sink or swim together?

Now that we've taken care of splitting passenger name into some new variables, we can take it a step further and make some new family variables. First we're going to make a **family size** variable based on number of siblings/spouse(s) (maybe someone has more than one spouse?) and number of children/parents. 

```{r}
# Create a family size variable including the passenger themselves
full$Fsize <- full$SibSp + full$Parch + 1

# Create a family variable 
full$Family <- paste(full$Surname, full$Fsize, sep='_')
```

What does our family size variable look like? To help us understand how it may relate to survival, let's plot it among the training data.

```{r, message=FALSE, warning=FALSE}
# Use ggplot2 to visualize the relationship between family size & survival
ggplot(full[1:891,], aes(x = Fsize, fill = factor(Survived))) +
  geom_bar(stat='count', position='dodge') +
  scale_x_continuous(breaks=c(1:11)) +
  labs(x = 'Family Size') +
  theme_few()
```

Ah hah. We can see that there's a survival penalty to singletons and those with family sizes above 4. We can collapse this variable into three levels which will be helpful since there are comparatively fewer large families. Let's create a **discretized family size** variable.

```{r}
# Discretize family size
full$FsizeD[full$Fsize == 1] <- 'singleton'
full$FsizeD[full$Fsize < 5 & full$Fsize > 1] <- 'small'
full$FsizeD[full$Fsize > 4] <- 'large'

# Show family size by survival using a mosaic plot
mosaicplot(table(full$FsizeD, full$Survived), main='Family Size by Survival', shade=TRUE)
```

The mosaic plot shows that we preserve our rule that there's a survival penalty among singletons and large families, but a benefit for passengers in small families. I want to do something further with our age variable, but `r sum(is.na(full$Age))` rows have missing age values, so we will have to wait until after we address missingness.

## Treat a few more variables ...

What's left? There's probably some potentially useful information in the **passenger cabin** variable including about their **deck**. Let's take a look.

```{r}
# This variable appears to have a lot of missing values
full$Cabin[1:28]

# The first character is the deck. For example:
strsplit(full$Cabin[2], NULL)
