---
title: "Titanic Survival Prediction"
author: "Courtney Wilkins"
date: May 14, 2017
output:
  html_document:
    toc: true
    toc_depth: 2
---

### In Progress ###

## Kaggle Competition Description

The sinking of the RMS Titanic is one of the most infamous shipwrecks in history. On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships.

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

In this challenge, we ask you to complete the analysis of what sorts of people were likely to survive. In particular, we ask you to apply the tools of machine learning to predict which passengers survived the tragedy.


1.  Set File Locations
---
Set all machine locations in one place so they can be easily found later
```{r locations}
# Input files 
in.train <- '../input/train.csv'
in.test <- '../input/test.csv'
```

2. Load Libraries
---
```{r libraries}
# Load libraries
library(readr) # CSV file I/O
library(ggplot2) # Data visualization

# Print out session data for reproducibility standards
sessionInfo()
```

3.  Load Data
---
Load Data File
```{r data}
# Read in data files
train <- read.csv(in.train, header = TRUE, row.names = 1)
test <- read.csv(in.test, header = TRUE, row.names = 1)
```

Input data files are available in the "../input/" directory.

For example, running this (by clicking run or pressing Shift+Enter) will list the files in the input directory

```{r}
system("ls ../input", intern=TRUE)
```

Any results you write to the current directory are saved as output.