---
title: "Surviving The Titanic"
author: "Lance J. Fernando"
output: html_document
---

This project is hosted by [Kaggle](https://www.kaggle.com) as an introduction to data mining and supervised learning. The data used is factual and is taken directly from their website. It contains all passengers that were aboard on the Titanic in April of 1912. Attributes for each passenger includes their Age, Gender, Fare price, etc. 

In this project we will be predicting the passenger's **Survival** based on these various attributes. First we will conduct some data preprocessing and analysis. Then will fit models using *logistic regression* and *randomForests* to drive our predictions.

Here is a short summary of our dataset.
```{r, message=FALSE, echo = FALSE}
library(randomForest)
tit_train <- read.csv("../input/train.csv", header = TRUE)
tit_test <- read.csv("../input/test.csv", header = TRUE)
head(tit_train)
```

Notice that all passengers have a title after their surname. In order to extract this, lets split up each name observation using [,.] as our delimeter. With this we can extract titles.
```{r}
names.split <- strsplit(as.character(tit_train$Name), "[,.]")
test.names.split <- strsplit(as.character(tit_test$Name), "[,.]")

title <- rep(NA, length(names.split))
test.title <- rep(NA, length(test.names.split))

for(i in 1:length(names.split)){
  title[i] <- trimws(names.split[[i]][2])
}

for(i in 1:length(test.names.split)){
  test.title[i] <- trimws(test.names.split[[i]][2])
}
table(title)
table(test.title)
```

Now with a bit of google searching we can extract titles that are of higher class. We can also translate foreign titles like "Mlle" and "Mme" and change them to "Mrs" and "Ms" respectively. We will create a vector of titles that grant more respect and create a dummy variable in our dataset. 
```{r}
respect <- c("Capt", "Col", "Don", "Dona",  "Dr", "Jonkheer", "Lady", "Major", "Rev", "Sir", "the Countess")

tit_train$Title <- title
tit_test$Title <- test.title

tit_train$Title[which(tit_train$Title == "Mlle")] <- "Ms"
tit_test$Title[which(tit_test$Title == "Mlle")] <- "Ms"

tit_train$Title[which(tit_train$Title == "Mme")] <- "Mrs"
tit_test$Title[which(tit_test$Title == "Mme")] <- "Mrs"

tit_train$Title[which(tit_train$Title == "Don")] <- "Sir"
tit_test$Title[which(tit_test$Title == "Don")] <- "Sir"

tit_train$Title[which(tit_train$Title == "Dona")] <- "Lady"
tit_test$Title[which(tit_test$Title == "Dona")] <- "Lady"

tit_test$Title[which(tit_test$Title %in% respect)] <- "Resp"
tit_train$Title[which(tit_train$Title %in% respect)] <- "Resp"

tit_train$Title <- factor(tit_train$Title)
tit_test$Title <- factor(tit_test$Title)

levels(tit_test$Title) <- levels(tit_train$Title)
```

Lets anaylze the survival rate based on Sex.
```{r}
table(Survived = tit_train$Survived, Sex = tit_train$Sex)
barplot(table(tit_train$Survived, tit_train$Sex), legend = c('Died', 'Survived'),
        xlab = 'Sex', ylab = 'Frequency', main = 'Female vs Male Survival Rate',
        col = c("darkred", "lightblue"))
```

Age is an important feature to include however many observations are missing from that column. In order to extract the essence of age out of each passenger, we can use their title along with other features.

Since mostly women survived, lets anaylze which among those women were more likely to make it out.
```{r}
table(Survived = tit_train$Survived[tit_train$Sex == 'female'], title[tit_train$Sex == 'female'])
```

Lets change the Sex values to 0 and 1 for female and male respectively.
```{r}
tit_train$Sex <- ifelse(tit_train$Sex == "male", 1, 0)
tit_test$Sex <- ifelse(tit_test$Sex == "male", 1, 0)
```
Women with titles 'Miss' and 'Mrs' occur more often so lets dive further into that data.

We can easily categorize women with the title of 'Mrs' to be married so lets make that into a feature. In addition, the title of 'Master' for men means that they are under the age of 18. Lets also create that feature.
```{r}
marrWom <- rep(0, nrow(tit_train))
marrWom.test <- rep(0,nrow(tit_test))
marrWom <- ifelse(title == "Mrs", 1, 0)
marrWom.test <- ifelse(test.title == "Mrs", 1, 0)
tit_train$marrWom <- marrWom
tit_test$marrWom <- marrWom.test

isBoy <- rep(0, nrow(tit_train))
isBoy.test <- rep(0,nrow(tit_test))
isBoy <- ifelse(title == "Master", 1, 0)
isBoy.test <- ifelse(test.title == "Master", 1, 0)

tit_train$isBoy <- isBoy
tit_test$isBoy <- isBoy.test
```

Lets analyze the age range for those with the title of 'Miss' or 'Ms'. We can also include the column regarding the number of Siblings/Spouses and Parents/Children.
```{r}
barplot(table(tit_train$Parch[title=="Miss"],tit_train$Age[title == "Miss"]),
        legend = c("0 Parents/Children", "1 Parent/Child", "2 Parents/Children"),
        xlab = "Age", ylab = "Frequency", main = "Ages of those titled 'Miss'")

barplot(table(tit_train$SibSp[title=="Miss"],tit_train$Age[title == "Miss"]),
        legend = c("0 Siblings/Spouses", "1 Sib/Spouse", "2 Sib/Spouse",
                   "3 Sib/Spouse", "4 Sib/Spouse", "5 Sib/Spouse", "8 Sib/Spouse"),
        col = c("darkslategrey", "darkslategray4", "darkslategray3", "darkslategray1",
                "goldenrod4", "goldenrod3", "goldenrod1"),
        xlab = "Age", ylab = "Frequency", main = "Ages of those titled 'Miss'")
mean(tit_train$Age[title == 'Miss' | title == 'Ms'], na.rm = TRUE)
sqrt(var(c(tit_train$Age[title == 'Miss' | title == 'Ms'], tit_test$Age[test.title=='Miss'|test.title=='Ms']), na.rm = TRUE))

```

The average age for 'Miss'/'Ms' is 21.8 with a standard deviation of 12.2. Lets take the threshold for a young female's age to be .25 deviations from the mean, giving us an age of about 25. In addition, those that are older tend to have smaller family sizes; and more importantly travel alone. So using this information we can create another feature that will categorize 'Miss'/'Ms' whether they are under or over this mean age.
```{r}
youngFem <- rep(0, nrow(tit_train))
youngFem.test <- rep(0,nrow(tit_test))
miss <- c(which(title == 'Miss'), which(title == 'Ms'))
miss.test <- c(which(test.title == 'Miss'), which(test.title == 'Ms'))
for(i in miss){
  if(is.na(tit_train$Age[i])){
    if(tit_train$SibSp[i] >= 1 || tit_train$Parch[i] >= 1)
      youngFem[i] <- 1
  }
  else if(tit_train$Age[i] <= 25)
    youngFem[i] <- 1
}
for(i in miss.test){
  if(is.na(tit_test$Age[i])){
    if(tit_test$SibSp[i] >= 1 || tit_test$Parch[i] >= 1)
      youngFem.test[i] <- 1
  }
  else if(tit_test$Age[i] <= 25)
    youngFem.test[i] <- 1
}

tit_train$youngFem <- youngFem
tit_test$youngFem <- youngFem.test
```

Lets also add a variable for family size. We add 1 to account for the passengers themselves as well.
```{r}
tit_train$famSize <- tit_train$SibSp + tit_train$Parch + 1
tit_test$famSize <- tit_test$SibSp + tit_test$Parch + 1
```

In the test set there seems to be an NA value for Fare. The passenger is of Pclass 3. Lets analyze how Pclass and Fare are correlated and give that passenger the mean Fare price for those of Pclass 3.
```{r}
missing.Fare <- which(is.na(tit_test$Fare))
tit_test[missing.Fare,]
cor(tit_train$Fare, tit_train$Pclass)
tit_test$Fare[missing.Fare] <- mean(tit_train$Fare[tit_train$Pclass == 3])
```


Lets take a look at the **Fare** variable. Here is a plot of its density. As you can see it is highly skewed so we will use a log transformation to fix that.
The second graph is much easier to interpret and use for modeling. We will go ahead and change that in our dataset.
```{r, echo = FALSE}
par(mfrow =c(2,2))
plot(density(tit_train$Fare), main = "Train Fare")
plot(density(log(tit_train$Fare)), main = "Train Log(Fare)")
plot(density(tit_test$Fare), main = "Test Fare")
plot(density(log(tit_test$Fare)), main = "Test Log(Fare)")
```

```{r}
dev.off()
tit_train$Fare <- log(1+tit_train$Fare)
tit_test$Fare <- log(1+tit_test$Fare)
```

Where people embarked from may turn out to be a significant feature. Analyzing the levels there are two passengers that have missing embark values. With a simple google search, these passengers boarded the Titanic at Southampton or 'S'. Lets just input that into our dataset
```{r}
levels(tit_train$Embarked)
levels(tit_test$Embarked)
missing.Embarked <- which(tit_train$Embarked == "")
tit_train$Embarked[missing.Embarked] <- "S"
tit_train[missing.Embarked,]
```
 
Now lets analyze the survival rate of each boarding station and create a new feature for each one.
```{r}
table(tit_train$Survived, tit_train$Embarked)
tit_train$Embarked.S <- ifelse(tit_train$Embarked == 'S', 1, 0)
tit_train$Embarked.Q <- ifelse(tit_train$Embarked == 'Q', 1, 0)
tit_train$Embarked.C <- ifelse(tit_train$Embarked == 'C', 1, 0)

tit_test$Embarked.S <- ifelse(tit_test$Embarked == 'S', 1, 0)
tit_test$Embarked.Q <- ifelse(tit_test$Embarked == 'Q', 1, 0)
tit_test$Embarked.C <- ifelse(tit_test$Embarked == 'C', 1, 0)


```

Lets also take a look at the cabin feature. This may or may not be important in our final model but it will not hurt to extract info out of it. We will create a feature for each cabin. Those that do not have their cabin labeled explicitly will get a 'U' for unknown.
```{r}
tit_train$cab <- ifelse(tit_train$Cabin == "" | tit_train$Cabin == "T", "U", substr(tit_train$Cabin,1,1))
tit_test$cab <- ifelse(tit_test$Cabin == "" | tit_test$Cabin == "T", "U", substr(tit_test$Cabin,1,1))

table(tit_train$Survived, tit_train$cab)

tit_train$cab.A <- ifelse(tit_train$cab == 'A', 1, 0)
tit_train$cab.B <- ifelse(tit_train$cab == 'B', 1, 0)
tit_train$cab.C <- ifelse(tit_train$cab == 'C', 1, 0)
tit_train$cab.D <- ifelse(tit_train$cab == 'D', 1, 0)
tit_train$cab.E <- ifelse(tit_train$cab == 'E', 1, 0)
tit_train$cab.F <- ifelse(tit_train$cab == 'F', 1, 0)
tit_train$cab.G <- ifelse(tit_train$cab == 'G', 1, 0)
tit_train$cab.U <- ifelse(tit_train$cab == 'U', 1, 0)

tit_test$cab.A <- ifelse(tit_test$cab == 'A', 1, 0)
tit_test$cab.B <- ifelse(tit_test$cab == 'B', 1, 0)
tit_test$cab.C <- ifelse(tit_test$cab == 'C', 1, 0)
tit_test$cab.D <- ifelse(tit_test$cab == 'D', 1, 0)
tit_test$cab.E <- ifelse(tit_test$cab == 'E', 1, 0)
tit_test$cab.F <- ifelse(tit_test$cab == 'F', 1, 0)
tit_test$cab.G <- ifelse(tit_test$cab == 'G', 1, 0)
tit_test$cab.U <- ifelse(tit_test$cab == 'U', 1, 0)
```

Now that we have an abundance of features lets start modeling!

We will begin with logistic regression and validate our misclassification error rate using cross validation. In order to show how to script cross validation lets try it with a simple model using Pclass and Sex. The success rate is printed below. Remember we do not want to overfit our model on the training set because that would lead to a reduction of accuracy on the test set.
```{r}
attach(tit_train)

K = 5
folds <- sample(1:K, nrow(tit_train), replace = TRUE)
error <- rep(0, 5)

for(i in 1:K){
  log.fit <- glm(factor(Survived)~Pclass + Sex,
                 data = tit_train,
                 subset = which(folds != i),
                 family = binomial)
  log.train.probs <- predict(log.fit, newdata = tit_train[folds == i,], type = "response")
  log.train.preds <- ifelse(log.train.probs >= 0.5, 1, 0)
  error[i] <- mean(log.train.preds != Survived[folds == i])
}
print(paste('Error rate: ', mean(error)))
print(paste('Success rate: ', 1-mean(error)))

```

Now lets create a model using random forests. This is the model used to score 0.80383 and landed me at number 1076 on the leaderboard as of Tue, 03 Jan 2017 22:06:22. Notice that it does not utilize all the features we had created initially.
```{r}
set.seed(1)
rf.fit <- randomForest(factor(Survived) ~Pclass + Sex + famSize+
                         isBoy + youngFem + marrWom 
                       + Embarked.S + Embarked.C + Embarked.Q,
                       data = tit_train, mtry = 2, ntree = 15000, nodesize = 1,
                       importance = FALSE)
rf.pred.train <- predict(rf.fit, newdata = tit_train, type = "class")
print(paste('Success rate on training data: ', mean(rf.pred.train == tit_train$Survived)))

rf.pred.out <- predict(rf.fit, newdata = tit_test, type = "class")
```

```{r, echo = FALSE}
results <- data.frame(PassengerId = 892:1309, Survived = rf.pred.out)
write.csv(results,"submission.csv", row.names =FALSE)

```