---
title: 'Exploring the Titanic Dataset'
author: 'Megan L. Risdal / modified by antagomir'
date: '27 Feb 2017'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is my first stab at a Kaggle script. I have chosen to work with the Titanic dataset after spending some time poking around on the site and looking at other scripts made by other Kagglers for inspiration. I will also focus on doing some illustrative data visualizations along the way. I'll then use `randomForest` to create a model predicting survival on the Titanic. I am new to machine learning and hoping to learn a lot, so feedback is very welcome! 

There are three parts to my script as follows:

* Feature engineering
* Missing value imputation
* Prediction!

## Load and check data

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('randomForest') # classification algorithm
```

Now that our packages are loaded, let's read in and take a peek at the data.

```{r, message=FALSE, warning=FALSE}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)

full  <- bind_rows(train, test) # bind training & test data

# check data
str(full)
```

We've got a sense of our variables, their class type, and the first few observations of each. We know we're working with 1309 observations of 12 variables. To make things a bit more explicit since a couple of the variable names aren't 100% illuminating, here's what we've got to deal with:

Variable Name | Description
--------------|-------------
Survived      | Survived (1) or died (0)
Pclass        | Passenger's class
Name          | Passenger's name
Sex           | Passenger's sex
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation


# Feature Engineering
## What's in a name?

The first variable which catches my attention is **passenger name** because we can break it down into additional meaningful variables which can feed predictions or be used in the creation of additional new variables. For instance, **passenger title** is contained within the passenger name variable and we can use **surname** to represent families. Let's do some **feature engineering**!

```{r, message=FALSE, warning=FALSE}
# Grab title from passenger names
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Show title counts by sex
table(full$Sex, full$Title)

# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
full$Title[full$Title == 'Mlle']        <- 'Miss' 
full$Title[full$Title == 'Ms']          <- 'Miss'
full$Title[full$Title == 'Mme']         <- 'Mrs' 
full$Title[full$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full$Sex, full$Title)

# Finally, grab surname from passenger name
full$Surname <- sapply(full$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])
```

```{r results='asis'}
cat(paste('We have <b>', nlevels(factor(full$Surname)), '</b> unique surnames. I would be interested to infer ethnicity based on surname --- another time.'))
```


---