---
title: 'Exploring the Titanic Dataset'
author: 'zossimov'
date: '21 November 2016'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is a first stab at a Kaggle script and a novice's first effort; forked from Megan Risdal's excellent one!
Based on DataCamp tutorial.

## Load and check data

```{r, message = FALSE}
# Load packages
library(rpart)
library(dplyr)

```

Now that our packages are loaded, let's read in and take a peek at the data.

```{r, message=FALSE, warning=FALSE}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)
full  <- bind_rows(train, test) # bind training & test data

```

We've got a sense of our variables, their class type, and the first few observations of each. We know we're working with 1309 observations of 12 variables. To make things a bit more explicit since a couple of the variable names aren't 100% illuminating, here's what we've got to deal with:

Variable Name | Description
--------------|-------------
Survived      | Survived (1) or died (0)
Pclass        | Passenger's class
Name          | Passenger's name
Sex           | Passenger's sex
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation


## Build a Model
We then build our model using `rpart` on the training set.

```{r}

# Build the model (note: not all possible variables are used)
# Build the decision tree
tree_model <- rpart(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked, data = train, method = "class")

```
## Alternative Model

```{r}
#Titles Grab title from passenger names
full$Title <- gsub('(.*, )|(\\..*)', '', full$Name)

# Show title counts by sex
table(full$Sex, full$Title)

# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
full$Title[full$Title == 'Mlle']        <- 'Miss' 
full$Title[full$Title == 'Ms']          <- 'Miss'
full$Title[full$Title == 'Mme']         <- 'Mrs' 
full$Title[full$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(full$Sex, full$Title)

full01 <- full
full01$Sex <- as.factor(full01$Sex)
full01$Embarked <- as.factor(full01$Embarked)
full01$Title <- as.factor(full01$Title)

train_new <- full01[full01$PassengerId %in% train$PassengerId,]
str(train_new)

test_new <- full01[full01$PassengerId %in% test$PassengerId,]
str(test_new)

# Finish the command
my_tree_five <- rpart(Survived ~ Pclass + Sex + Age + SibSp + 
                        Parch + Fare + Embarked + Title,
                      data = train_new, method = "class")

```

## Prediction!

```{r}

# Make predictions on the test set
my_prediction <- predict(tree_model, newdata = test, type = "class")

# Finish the data.frame() call
my_solution <- data.frame(PassengerId = test$PassengerId, Survived = my_prediction)

# Use nrow() on my_solution to validate format is right
nrow(my_solution);ncol(my_solution)

# Write the solution to file
write.csv(my_solution, file = 'tree_basic_Solution.csv', row.names = FALSE)

# Make prediction with alt model
my_prediction <- predict(my_tree_five, test_new, type = "class")

# Make results ready for submission
my_solution <- data.frame(PassengerId = test_new$PassengerId, Survived = my_prediction)

# Write the solution to file
write.csv(my_solution, file = 'tree_title_Solution.csv', row.names = FALSE)


```


---