---
title: "Titanic survivor prediction"
author: "Ravi Krishna Reddy"
output: 
  html_document: 
    number_sections: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
options(width = 100)
```

#Introduction
This is my first attempt in doing analysis and participating in Kaggle competition. 
I found Titanic dataset has very few variables which makes learning and working easy for a new bee like me. 
Initially I don't know where to start and how to approach. So I have followed Erik Bruin's kernal (who also inspired me a lot) and learned how to think in right direction. 
Much of the code here is just a copy from his kernal, except few exploratory data analysis steps and feature engineering which I have done on my own. Will try to articulate each and every step I did later.
A big thank you [Erik](https://www.kaggle.com/erikbruin). 


#Loading and Exploring Data

##Loading required libraries
```{r echo=FALSE}
library(ggplot2)
library(gridExtra)
library(dplyr)
library(forcats)
library(Hmisc)
library(randomForest)
library(caret)
library(corrplot)
```

##Loading data into R
```{r}
train <- read.csv("../input/train.csv", stringsAsFactors = FALSE, na.strings = c("NA", ""))
test <- read.csv("../input/test.csv", stringsAsFactors = FALSE, na.strings = c("NA", ""))
```

##Data size and structure
Dimensions and overview of train data
```{r}
dim(train)
head(train)
```

Dimensions and overview of test data
```{r}
dim(test)
head(test)
```

Observe that our train dataset has 12 variables whereas test dataset has 11 variables. This make sense because unlike train dataset, test dataset don't have the 'Survived' variable.
To make our data cleaning work easy and time saving, let us combine both the datasets into a new dataset 'all'. To combine two datasets into one, both of them must have equal number of variables, hence create a 'Survived' column with NAs in test dataset and them combine it with train dataset to create all dataset.

##Combining train and test data
```{r}
test$Survived <- NA
all <- rbind(train, test)
dim(all)
```

##Identifying variables with NA values
```{r}
sort(colSums(is.na(all)), decreasing = TRUE)
```

**Key observations:**  

* `Survived` has 418 NAs. These are the values we need to predict and hence we can ignore NAs in `Survived`.  
* Ignoring `Survived`, there are 4 variables with NAs.  
* `Cabin` is sparsily populated and has 1014 NAs.  
* `Age` has 263 NAs.  
* `Embarked` has 2 NAs.  
* `Fare` has 1 NA.

**Converting `Survived`, `Sex` and `Pclass` into factors:**
Since our response variable `Survived` is complete and tidy, let us convert it into a factor.  
Also, let us convert `Sex` and `Pclass` into factor as they are also complete and tidy.
```{r}
all$Survived <- as.factor(all$Survived)
all$Pclass <- as.ordered(all$Pclass) #Pclass has ordered values, so converting it into ordinal factor
all$Sex <- as.factor(all$Sex)
```

###Dealing with NAs in Embarked variable
Embarked details are missing for two passengers. A quick internet search reveals that they have embarked at Southempton. Hence, let's directly impute the missing values with 'S'

```{r}
all$Embarked[is.na(all$Embarked)] <- 'S'
all$Embarked <- as.factor(all$Embarked)
```

###Dealing with NAs in Fare variable
Fare is missing for one passenger with ticket number 3701. The ticket numbers before and after 3701 have the fare value 8.0500 and there is high chance that 3701 might also have same fare.
```{r}
all$Fare[is.na(all$Fare)] <- 8.0500
```


#Exploring some of the important variables
Let us now do some quick analysis on the data to understand more about the response variable and how the predictors are related to it.

##Response variable: Survived
```{r fig.align="center"}
ggplot(data=all[!is.na(all$Survived),], aes(x=Survived)) + 
  geom_bar(stat='count', aes(fill=Survived)) +
  geom_label(stat='count', aes(label=..count..)) + 
  labs(x="not survived and survived in train data") + 
  theme_grey(base_size = 15)
```

From above plot, it is clear that the count of people dead is more than survived.

##Predictor variable: Sex
```{r fig.align="center", fig.width=10}
p1 <- ggplot(data=all, aes(x=Sex)) + 
        geom_bar(stat='count', aes(fill=Sex)) + 
        geom_label(stat='count', aes(label=..count..)) + 
        labs(x="Gender ratio (total data)")

p2 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Sex, group=Survived)) + 
        geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
        geom_label(stat='count', aes(label=..count..), size = 3, position = position_dodge(0.9)) + 
        labs(x="Survived ratio by Gender") 

p3 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Sex, fill=Survived)) + 
        geom_bar(stat='count', position = "fill") + 
        labs(x="Survived ratio by Gender", y="percent")

grid.arrange(p1, p2, p3, ncol=3)
```

**Key observations:**  

* There are more male passengers than female passengers.
    + Nearly 64% male passengers and 36% female passengers.
* The survival chances of female are very high compared to male.
    + Among female passengers, 74% are survived and among male passengers, only 19% are survived.  

From these observations, it is clear that `Sex` is an important predictor.

##Predictor variable: Pclass
```{r fig.width=10, fig.align="center"}
p4 <- ggplot(data=all, aes(x=Pclass)) + 
        geom_bar(stat='count', aes(fill=Pclass), position="dodge") + 
        geom_label(stat="count", aes(label=..count..)) + 
        labs(x="Passenger class ratio (total data)")

p5 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass, group = Survived)) + 
        geom_bar(stat = "count", aes(fill=Survived), position="dodge") + 
        geom_label(stat = "count", aes(label=..count..), size=3, position = position_dodge(0.9)) + 
        labs(x="Survived ratio by Pclass")

p6 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass, fill=Survived)) + 
        geom_bar(stat="count", position="fill") + 
        labs(x="Survived ratio by Pclass", y="percent")

grid.arrange(p4, p5, p6, ncol=3)

p7 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass, group = Survived)) + 
        geom_bar(stat = "count", aes(fill=Survived), position="dodge") + 
        geom_label(stat = "count", aes(label=..count..), size=3, position = position_dodge(0.9)) + 
        labs(x="Survived ratio by Pclass and Gender") + 
        facet_grid(.~Sex)

p8 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass, fill=Survived)) + 
        geom_bar(stat = "count", position="fill") + 
        labs(x="Survived ratio by Pclass and Gender", y="Percent") + 
        facet_grid(.~Sex)

grid.arrange(p7, p8, ncol=2)
```

**Key observations:**  

* More passengers travelled in 3rd class.
    + Approximately, 25% travelled in 1st class, 21% travelled in 2nd class and 54% travelled in 3rd class.
* Passengers belonging to 3rd class have very less chances of surviving.
    + Approximately, 63% of 1st class, 47% of 2nd class and 24% of 3rd class passengers are survived.
* Female belonging to 1st and 2nd classes have very high chances of surviving.
    + Approximately, 97% of 1st class, 92% of 2nd class and 50% of 3rd class female passengers are survived
* The chances of surviving of male are almost very less irrespective of classes.
    + Approximately, 37% of 1st class, 16% of 2nd class and 14% of rd class male passengers are survived.  

From these observations, it is clear that `Pclass` is an important predictor.  

##Predictor variable: Embarked
**Embarked variable and it's relationship with response variable**
```{r fig.align="center", fig.width=10}
p9 <- ggplot(data=all[!is.na(all$Survived),], aes(x=fct_infreq(Embarked))) + 
        geom_bar(stat='count', aes(fill=Embarked)) + 
        geom_label(stat='count', aes(label=..count..)) + 
        labs(x="Passengers by Embarked")

p10 <- ggplot(data=all[!is.na(all$Survived),], aes(x=fct_infreq(Embarked), group=Survived)) + 
        geom_bar(stat='count', aes(fill=Survived), position="dodge") + 
        geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
       labs(x="Survived ratio by Embarked")

grid.arrange(p9, p10, ncol=2)
```

##Predictor variable: Fare
From a quick Google search, below are my observations regarding ticket fares.

* First Class (parlor suite) £870
* First Class (berth) £30
* Second Class £12
* Third Class £3 to £8

But surprisingly, fares in our data are very high for many passengers. In my investigation, I found that, the fare is not always fare per person. If tickets are booked as a group, the group price is being assigned to each passenger.  

I will work on this later in Feature Engineering section by creating a new variable `FarePP` (Fare Per Person).

#Feature Engineering  

##Creating GenderClass variable
If we have a close look at above analysis (Sex and Pclass), we can easily understand that the combination of `Sex` and `Pclass` is explaining more about response variable rather than individually. For example, Not all female have equal chances of surviving. Female belonging to 1st and 2nd classes have high chances compared to 3rd class female.  

So creating a new variable `GenderClass` (combination of Sex and Pclass) might really help. Let's do that!  

```{r}
all$GenderClass <- paste0(all$Sex, "P", all$Pclass)
all$GenderClass <- as.factor(all$GenderClass)
```

```{r fig.align="center", fig.width=10}
p9 <- ggplot(data = all[!is.na(all$Survived),], aes(x=GenderClass, group=Survived)) + 
  geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
  geom_label(stat='count', aes(label=..count..), size=3, position = position_dodge(0.9)) + 
  labs(x="Survived ratio by GenderClass")  

p10 <- ggplot(data = all[!is.na(all$Survived), ], aes(x=GenderClass)) + 
  geom_bar(stat = 'count', aes(fill=Survived), position = "fill") + 
  labs(x="Survived ratio by GenderClass", y="percent")

grid.arrange(p9, p10, ncol=2)
```


Let's create some new variables out of `Name` variable and check if they are capable of explaining the variation in response variable.  

##Creating Title variable 
```{r}
#760 "Rothes, Mrs. Norman Leslie (Lucy Noel Martha Dyer-Edwards)"
#711 "Mayne, Mlle. Berthe Antonine"
#797 "Leader, Dr. Alice (Farnham)"

all$Title <- sapply(all$Name, function(x){trimws(unlist(strsplit(x, split = '[,.]'))[2])})
table(all$Sex, all$Title)
```

Mr, Don, Sir, Jonkheer all mean Mister (Mr)
Mrs, Lady, the Countess all mean Missus (Mrs)
Miss, Ms, Dona, Mlle, Mme all mean Miss (Miss)  

Also, Rev, Major, Col and Capt being male here, we can impute them with the title Mr.
All the people with title Dr are male except one. For all male doctors impute title as Mr and for the one female doctor impute title as Mrs as she is married.
```{r}
all$Title[all$Title %in% c("Don", "Sir", "Jonkheer", "Rev", "Major", "Col", "Capt")] <- "Mr"
all$Title[all$Title %in% c("Lady", "the Countess")] <- "Mrs"
all$Title[all$Title %in% c("Ms", "Dona", "Mlle", "Mme")] <- "Miss"
all$Title[all$Title == 'Dr' & all$Sex == 'male'] <- "Mr"
all$Title[all$Title == 'Dr' & all$Sex == 'female'] <- "Mrs"

table(all$Sex, all$Title)
```

```{r fig.align="center", fig.width=10, fig.height=10}
p11 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass, group=Survived)) + 
        geom_bar(stat='count', aes(fill=Survived), position="dodge") + 
        geom_label(stat='count', aes(label=..count..), size=3, position = position_dodge(0.9)) + 
        labs(x="Survived ratio by Pclass and Title") + 
        facet_grid(.~Title)

p12 <- ggplot(data=all[!is.na(all$Survived),], aes(x=Pclass)) + 
        geom_bar(stat='count', aes(fill=Survived), position="fill") + 
        labs(x="Survived ratio by Pclass and Title", y="percent") + 
        facet_grid(.~Title)

grid.arrange(p11, p12, nrow=2)
```


##Creating TitleClass variable
```{r fig.align="center", fig.width=10}
all$TitleClass <- paste0(all$Title, "P", all$Pclass)
all$TitleClass <- as.factor(all$TitleClass)

p13 <-ggplot(data=all[!is.na(all$Survived),], aes(x=TitleClass)) + 
        geom_bar(stat='count', aes(fill=Survived), position="fill") + 
        labs(x="Survived ratio by TitleClass", y="percent")

grid.arrange(p13)
```

##Creating TicketPrefix
```{r}
all$TicketPrefix <- sapply(all$Ticket, function(x){trimws(unlist(strsplit(x, split=' '))[1])})
HasTicketPrefix <- sapply(all$Ticket, function(x){grepl("\\D", x)})
all$TicketPrefix[!HasTicketPrefix] <- "No Prefix"
all$TicketPrefix[all$TicketPrefix %in% c("PC")] <- "PC"
all$TicketPrefix[!all$TicketPrefix %in% c("PC")] <- "Other"
rm(HasTicketPrefix)
```

```{r fig.width=10, fig.align="center"}
p14 <- ggplot(data=all[!is.na(all$Survived),], aes(x=TicketPrefix, group=Survived)) + 
  geom_bar(stat='count', aes(fill=Survived), position = 'dodge') + 
  geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
  facet_grid(.~Sex)

p15 <- ggplot(data=all[!is.na(all$Survived),], aes(x=TicketPrefix)) + 
  geom_bar(stat='count', aes(fill=Survived), position = 'fill') + 
  labs(y="percent")

grid.arrange(p14, p15, ncol=2)
```

##Creating FSize variable
```{r fig.align="center", fig.width=10, fig.height=10}
all$FSize <- all$SibSp + all$Parch + 1

p16 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(FSize), group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position="dodge") +
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ratio by Family Size") 

p17 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(FSize))) + 
          geom_bar(stat='count', aes(fill=Survived), position="fill") + 
          labs(x="Survived ratio by Family Size", y="percent") 

grid.arrange(p16, p17, nrow=2)
```

##Creating FSizeGroup
```{r}
GroupFunction <- function(x){
  if(x==1){
    return("Solo")
  }else if(x==2){
    return("Duo")
  }else if(x>=3 & x<=4){
    return("SmallFamily")
  }else{
    return("LargeFamily")
  }
}

all$FSizeGroup <- sapply(all$FSize, GroupFunction)
rm(GroupFunction)
```

```{r fig.align="center", fig.width=10}
p18 <- ggplot(data=all[!is.na(all$Survived),], aes(x=FSizeGroup, group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ratio by Family Group")

p19 <- ggplot(data=all[!is.na(all$Survived),], aes(x=FSizeGroup)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "fill") + 
          labs(x="Survived ratio by Family Group", y="percent")

grid.arrange(p18, p19, ncol=2)
```


##Creating PPTicket and TicketShared variables
```{r}
PPTicket <- all %>% group_by(Ticket=all$Ticket) %>% summarise(PPTicket=n())
all$PPTicket <- PPTicket$PPTicket[match(all$Ticket, PPTicket$Ticket)]
rm(PPTicket)

all$TicketShared <- sapply(all$PPTicket, function(x){ifelse(x==1, "Not Shared", "Shared")})
```


```{r fig.align="center", fig.width=10, fig.height=10}
p20 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(PPTicket), group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ratio by Passengers per Ticket")

p21 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(PPTicket))) + 
          geom_bar(stat='count', aes(fill=Survived), position = "fill") + 
          labs(x="Survived ratio by Passengers per Ticket", y="percent") 

grid.arrange(p20, p21, nrow=2)
```


```{r fig.align="center", fig.width=10}
p22 <- ggplot(data=all[!is.na(all$Survived),], aes(x=TicketShared, group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ratio by Sharing Ticket")

p23 <- ggplot(data=all[!is.na(all$Survived),], aes(x=TicketShared)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "fill") + 
          labs(x="Survived ratio by Sharing Ticket", y="percent") 

grid.arrange(p22, p23, ncol=2)
```


##Creating PGroupSize variable
```{r}
all$PGroupSize <- mapply(function(a, b){max(a, b)}, all$FSize, all$PPTicket)
all$PGroupSize <- as.numeric(all$PGroupSize)
```


##Creating PGroup variable
```{r}
GroupFunction <- function(x){
  if(x==1){
    return("Single")
  }else if(x==2){
    return("Duo")
  }else if(x>=3 & x<=4){
    return("Group")
  }else{
    return("Large Group")
  }
}

all$PGroup <- sapply(all$PGroupSize, GroupFunction)
rm(GroupFunction)
```


```{r fig.align="center", fig.width=10}
p24 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(PGroup), group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ratio by Passenger Group")

p25 <- ggplot(data=all[!is.na(all$Survived),], aes(x=factor(PGroup))) + 
          geom_bar(stat='count', aes(fill=Survived), position = "fill") + 
          labs(x="Survived ratio by Passenger Group", y="percent")

grid.arrange(p24, p25, ncol=2)
```


##Creating CabinPrefix variable
```{r}
CabinDetails <- all[!is.na(all$Cabin), c("Ticket", "Cabin")]
all$CabinModified <- CabinDetails$Cabin[match(all$Ticket, CabinDetails$Ticket)]
all$CabinModified[is.na(all$CabinModified)] <- "N"
all$CabinPrefix <- substr(all$CabinModified, 1, 1)

all$CabinModified <- NULL
rm(CabinDetails)
```

```{r fig.align="center", fig.width=10, fig.height=10}
p26 <- ggplot(data=all[!is.na(all$Survived),], aes(x=CabinPrefix, group=Survived)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9)) + 
          labs(x="Survived ration by Cabin Class")

p27 <- ggplot(data=all[!is.na(all$Survived),], aes(x=CabinPrefix)) + 
          geom_bar(stat='count', aes(fill=Survived), position = "fill") + 
          labs(x="Survived ration by Cabin Class", y="percent") 
grid.arrange(p26, p27, nrow=2)
```


##Creating FarePP variable
```{r}
all$FarePP <- all$Fare/all$PPTicket
```

```{r fig.width=10, fig.align=10}
ggplot(data=all[all$CabinPrefix!="N",], aes(x=CabinPrefix, group=Pclass)) + 
          geom_bar(stat='count', aes(fill=Pclass), position = "dodge") + 
          geom_label(stat='count', aes(label=..count..), position = position_dodge(0.9))
```

```{r}
ggplot(data=all, aes(x=FarePP)) + 
          geom_histogram(binwidth = 5, fill='blue', boundary=0) + 
          scale_x_continuous(breaks = seq(0, 150, by=5))
```

```{r}
all$FareBins <- cut2(all$FarePP, g=5)
```

##Predicting missing Ages
```{r}
ggplot(data=all[!is.na(all$Survived) & !is.na(all$Age),], aes(x=Age, group=Survived)) + 
  geom_density(alpha=0.5, aes(fill=Survived)) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n=10))
```

```{r}
AgeLM <- lm(Age ~ Pclass + Title + SibSp + Parch, data=all[!is.na(all$Age),])
summary(AgeLM)
```
```{r}
all$AgeLM <- predict(AgeLM, newdata = all)
```

#Setting datatypes of the variables
```{r}
all$Title <- as.factor(all$Title)
all$TicketPrefix <- as.factor(all$TicketPrefix)
all$FSizeGroup <- as.factor(all$FSizeGroup)
all$TicketShared <- as.factor(all$TicketShared)
all$PGroup <- as.factor(all$PGroup)
all$CabinPrefix <- as.factor(all$CabinPrefix)
```


#Model buildiing
```{r}
#quick_RF <- randomForest(x=all[!is.na(all$Survived),-c(1, 2, 4, 6, 9, 11)], y=all[!is.na(all$Survived),2], ntree=100, importance = TRUE)
#quick_RF <- randomForest(Survived ~ ., data=all[!is.na(all$Survived),-c(1, 4, 6, 9, 11)], ntree=1000, importance = TRUE)
#varImpPlot(quick_RF, main = "Variable Importance")
```

```{r}
trainClean <- all[!is.na(all$Survived), c("Survived", "TitleClass", "FarePP", "PGroupSize")]
testClean <- all[is.na(all$Survived), c("Survived", "TitleClass", "FarePP", "PGroupSize")]
```

##Random Forest
```{r}
set.seed(2018)
model_RF <- train(x=trainClean[,-1], y=trainClean$Survived, data=trainClean, method='rf', trControl=trainControl(method = 'cv', number=10))
model_RF
model_RF$results
```


```{r}
varImpPlot(model_RF$finalModel)
```

```{r}
prediction_RF <- predict(model_RF, testClean)
```


##Support Vector Machine
```{r}
set.seed(2018)
model_SVM <- train(Survived ~ TitleClass + FarePP + PGroupSize, data=trainClean, method='svmRadial', preProcess= c('center', 'scale'), trControl=trainControl(method="cv", number=10))
model_SVM
model_SVM$results
```

```{r}
prediction_SVM <- predict(model_SVM, testClean)
```


##Gradient Boosting Model
```{r}
set.seed(2018)
model_GBM <- train(Survived~ TitleClass + FarePP + PGroupSize, data=trainClean, method='gbm', preProcess= c('center', 'scale'), trControl=trainControl(method="cv", number=10), verbose=FALSE)
model_GBM
model_GBM$results
```

```{r}
prediction_GBM <- predict(model_GBM, testClean)
```

#Combining models
```{r}

testClean$RF <- as.numeric(prediction_RF)-1
testClean$SVM <- as.numeric(prediction_SVM)-1
testClean$GBM <- as.numeric(prediction_GBM)-1


corrplot.mixed(cor(testClean[, c('RF', 'SVM', 'GBM')]), order="hclust", tl.col="black")
```

```{r}
testClean$Sum <- testClean$RF + testClean$SVM + testClean$GBM
testClean$Majority <- ifelse(testClean$Sum<=1, 0, 1)
```

#Submission
```{r}
submission_select <- data.frame(PassengerId = test$PassengerId, Survived = testClean$Majority)
write.csv(submission_select, file='Titanic_Submission.csv', row.names = FALSE)
```

