---
title: 'Exploring the Titanic Dataset'
author: 'Megan L. Risdal'
date: '6 March 2016'
output:
  html_document:
    number_sections: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: readable
    highlight: tango
---

# Introduction

This is my first stab at a Kaggle script. I have chosen to work with the Titanic dataset after spending some time poking around on the site and looking at other scripts made by other Kagglers for inspiration. I will also focus on doing some illustrative data visualizations along the way. I'll then use `randomForest` to create a model predicting survival on the Titanic. I am new to machine learning and hoping to learn a lot, so feedback is very welcome! 

There are three parts to my script as follows:

* Feature engineering
* Missing value imputation
* Prediction!

## Load and check data

```{r, message = FALSE}
# Load packages
library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('dplyr') # data manipulation
library('mice') # imputation
library('randomForest') # classification algorithm
```

Now that our packages are loaded, let's read in and take a peek at the data.

```{r, message=FALSE, warning=FALSE}
train <- read.csv('../input/train.csv', stringsAsFactors = F)
test  <- read.csv('../input/test.csv', stringsAsFactors = F)

full  <- bind_rows(train, test) # bind training & test data

# check data
str(full)
```

We've got a sense of our variables, their class type, and the first few observations of each. We know we're working with 1309 observations of 12 variables. To make things a bit more explicit since a couple of the variable names aren't 100% illuminating, here's what we've got to deal with:

Variable Name | Description
--------------|-------------
Survived      | Survived (1) or died (0)
Pclass        | Passenger's class
Name          | Passenger's name
Sex           | Passenger's sex
Age           | Passenger's age
SibSp         | Number of siblings/spouses aboard
Parch         | Number of parents/children aboard
Ticket        | Ticket number
Fare          | Fare
Cabin         | Cabin
Embarked      | Port of embarkation


# Feature Engineering
## What's in a name?

